\documentclass{article}
\usepackage[backend=biber,natbib=true,style=alphabetic,maxbibnames=50]{biblatex}
\addbibresource{/home/nqbh/reference/bib.bib}
\usepackage[utf8]{vietnam}
\usepackage{tocloft}
\renewcommand{\cftsecleader}{\cftdotfill{\cftdotsep}}
\usepackage[colorlinks=true,linkcolor=blue,urlcolor=red,citecolor=magenta]{hyperref}
\usepackage{amsmath,amssymb,amsthm,enumitem,float,graphicx,mathtools,tikz}
\usetikzlibrary{angles,calc,intersections,matrix,patterns,quotes,shadings}
\allowdisplaybreaks
\newtheorem{assumption}{Assumption}
\newtheorem{baitoan}{}
\newtheorem{cauhoi}{Câu hỏi}
\newtheorem{conjecture}{Conjecture}
\newtheorem{corollary}{Corollary}
\newtheorem{dangtoan}{Dạng toán}
\newtheorem{definition}{Definition}
\newtheorem{dinhly}{Định lý}
\newtheorem{dinhnghia}{Định nghĩa}
\newtheorem{example}{Example}
\newtheorem{ghichu}{Ghi chú}
\newtheorem{hequa}{Hệ quả}
\newtheorem{hypothesis}{Hypothesis}
\newtheorem{lemma}{Lemma}
\newtheorem{luuy}{Lưu ý}
\newtheorem{nhanxet}{Nhận xét}
\newtheorem{notation}{Notation}
\newtheorem{note}{Note}
\newtheorem{principle}{Principle}
\newtheorem{problem}{Problem}
\newtheorem{proposition}{Proposition}
\newtheorem{question}{Question}
\newtheorem{remark}{Remark}
\newtheorem{theorem}{Theorem}
\newtheorem{vidu}{Ví dụ}
\usepackage[left=1cm,right=1cm,top=5mm,bottom=5mm,footskip=4mm]{geometry}
\def\labelitemii{$\circ$}
\DeclareRobustCommand{\divby}{%
	\mathrel{\vbox{\baselineskip.65ex\lineskiplimit0pt\hbox{.}\hbox{.}\hbox{.}}}%
}
\setlist[itemize]{leftmargin=*}
\setlist[enumerate]{leftmargin=*}

\title{Advanced STEM Students}
\author{Nguyễn Quản Bá Hồng\footnote{A scientist- {\it\&} creative artist wannabe, a mathematics {\it\&} computer science lecturer of Department of Artificial Intelligence {\it\&} Data Science (AIDS), School of Technology (SOT), UMT Trường Đại học Quản lý {\it\&} Công nghệ TP.HCM, Hồ Chí Minh City, Việt Nam.\\E-mail: {\sf nguyenquanbahong@gmail.com} {\it\&} {\sf hong.nguyenquanba@umt.edu.vn}. Website: \url{https://nqbh.github.io/}. GitHub: \url{https://github.com/NQBH}.}}
\date{\today}

\begin{document}
\maketitle
\begin{abstract}
	This text is a part of the series {\it Some Topics in Advanced STEM \& Beyond}:
	
	{\sc url}: \url{https://nqbh.github.io/advanced_STEM/}.
	
	Latest version:
	\begin{itemize}
		\item {\it Advanced STEM Students}.
		
		PDF: {\sc url}: \url{https://github.com/NQBH/advanced_STEM_beyond/blob/main/teach/student/NQBH_student.pdf}.
		
		\TeX: {\sc url}: \url{https://github.com/NQBH/advanced_STEM_beyond/blob/main/teach/student/NQBH_student.tex}.
		\item {\it }.
		
		PDF: {\sc url}: \url{.pdf}.
		
		\TeX: {\sc url}: \url{.tex}.
	\end{itemize}
\end{abstract}
\tableofcontents

%------------------------------------------------------------------------------%

\section{UMT Summer Semester 2025{\tt/}1480 + 1481: Introduction to Mathematical Analysis}

\begin{enumerate}
	\item {\sc Lê Minh Trâm Anh.}
	\begin{enumerate}
		\item Early attendance: 1.
		\item Bonus: 
	\end{enumerate}
	\item {\sc Nguyễn Bảo Anh.}
	\item {\sc Cao Gia Bảo.}
	\begin{itemize}
		\item Absence: 1.
	\end{itemize}
	\item {\sc Trần Văn Hậu.}
	\begin{itemize}
		\item Absence: 1.
	\end{itemize}
	\item {\sc Nguyễn Ngọc Bích Hiền.}
	\begin{enumerate}
		\item Early attendance: 1.
		\item Bonus: +0.25.
	\end{enumerate}
	\item {\sc Nguyễn Văn Hiếu.}
	\item {\sc Dương Xuân Hưng.}
	\item {\sc Trần Lê Gia Huy.}
	\item {\sc Trần Quang Huy.}
	\item {\sc Nguyễn Phúc An Khang.}
	\item {\sc Lương Quốc Khánh.}
	\item {\sc Nguyễn Hà Đăng Khoa}.
	\item {\sc Nguyễn Huy Đăng Khoa.}
	\item {\sc Nguyễn Lê Đăng Khoa.}
	\item {\sc Lê Bảo Ngọc.}
	\begin{enumerate}
		\item Early attendance: 1.
		\item Bonus: +0.25.
	\end{enumerate}
	\item {\sc Nguyễn Huỳnh Chí Lâm.}
	\item {\sc Nguyễn Lương Nghĩa.}
	\begin{enumerate}
		\item Early attendance: 1.
		\item Bonus: +0.5.
	\end{enumerate}
	\item {\sc Trần Phúc Nguyên.}
	\item {\sc Bùi Văn Nhớ}.
	\begin{itemize}
		\item +40 mins.
	\end{itemize}
	\item {\sc Lê Thanh Tân.}
	\item {\sc Kiều Thu.}
	\item {\sc Hoàng Nghĩa Tín.}
	\item {\sc Hoàng Anh Tuấn.} -1
	\item {\sc Ngô Thanh Tùng.}
	\item {\sc Ngô Hoàng Tùng.}
	\item {\sc Trần Mộng Tuyền.}
	\item {\sc Nguyễn Quang Vỹ.}	
\end{enumerate}

%------------------------------------------------------------------------------%

\section{UMT Summer Semester 2025{\tt/}1387: Combinatorics \& Graph Theory}

\begin{enumerate}
	\item {\sc Võ Ngọc Trâm Anh.}
	\item {\sc Hoàng Anh.}
	\item {\sc Sơn Tân.}
	\item {\sc Phan Vĩnh Tiến.}
\end{enumerate}

%------------------------------------------------------------------------------%

\section{UMT Summer Semester 2025{\tt/}1488: Introduction to AI}

%------------------------------------------------------------------------------%

\section{UMT Summer Semester 2025{\tt/}1493: Information Technology Fundamentals 2}

%------------------------------------------------------------------------------%

\section{Miscellaneous}

%------------------------------------------------------------------------------%

\printbibliography[heading=bibintoc]
	
\end{document}