\documentclass[onsided]{article}
\usepackage[backend=biber,natbib=true,style=authoryear]{biblatex}
\addbibresource{/home/hong/1_NQBH/math/bib.bib}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage[colorlinks=true,linkcolor=blue,urlcolor=red,citecolor=magenta]{hyperref}
\usepackage{amsmath,amssymb,amsthm}
\allowdisplaybreaks
\numberwithin{equation}{section}
\newtheorem{assumption}{Assumption}[section]
\newtheorem{lemma}{Lemma}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{definition}{Definition}[section]
\newtheorem{proposition}{Proposition}[section]
\newtheorem{theorem}{Theorem}[section]
\newtheorem{notation}{Notation}[section]
\newtheorem{remark}{Remark}[section]
\newtheorem{example}{Example}[section]
\newtheorem{ques}{Question}[section]
\newtheorem{problem}{Problem}[section]
\newtheorem{conjecture}{Conjecture}[section]
\usepackage[left=0.5in,right=0.5in,top=1.5cm,bottom=1.5cm]{geometry}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\addtolength{\headheight}{0pt}% obsolete
\lhead{\itshape \scriptsize \chaptername~\thechapter}
\rhead{\itshape \scriptsize \nouppercase{\leftmark}} %\nouppercase !
\renewcommand{\sectionmark}[1]{\markboth{#1}{}}
\cfoot{\thepage}

\title{Timeline of organizational and administrative aspects of PhD project}
\author{Nguyen Quan Ba Hong}
\date{\today}

\begin{document}
\maketitle
\setcounter{secnumdepth}{6}
\setcounter{tocdepth}{6}
\tableofcontents

%------------------------------------------------------------------------------%

\section{Timeline}
\begin{itemize}
    \item \textbf{Jul 20--24, 2020}: \href{https://conferences.leeds.ac.uk/lesgi/}{162nd European Study Group with Industry}, University of Leeds, UK.
    \item \textbf{Sep 7--10, 2020}: \href{https://morss2020.epfl.ch/}{Model Order Reduction Summer School 2020} (MORSS2020), \'Ecole polytechnique f\'ed\'erale de Lausanne (EPFL), Lausanne, Switzerland.
    \item \textbf{Sep 14--17, 2020}: \href{https://www.mathematik.de/}{Deutsche Mathematiker-Vereinigung} (German Mathematical Society) Annual Meeting 2020 (\href{https://www.tu-chemnitz.de/mathematik/dmv2020/index.php}{DMV2020}), Technische Universit\"at Chemnitz, Chemnitz, Germany.
    \item \textbf{Sep 15, 2020}: ROMSOC Ethical Monitoring - 3rd Workshop on Ethics.
    \item \textbf{Sep 28--29, 2020}: ROMSOC Internal Workshop.
    \item \textbf{Nov 30, 2020}: \href{https://www.romsoc.eu/blog/}{ROMSOC Blog}. Hong Nguyen, \href{https://www.romsoc.eu/optimal-shape-design-of-air-ducts/}{\textit{Optimal shape design of air ducts}} [\href{https://github.com/NQBH/miscellaneous/blob/master/ROMSOC/blog/optimal_shape_design_of_airducts/blog.pdf}{pdf}].
    \item \textbf{Nov 2020--Apr 2021}: ROMSOC Scientific Seminar.
    \item \textbf{Dec 18, 2020}: Industrial transfer report: \textit{Knowledge transfer to industrial partners and future valorization}.
    \item \textbf{Apr 1, 2021}: Secondment, Math.Tec, Vienna, Austria.
    \item \textbf{Apr 5, 2021}: ESR11: Hong Nguyen. Project 11: \textit{Optimal Shape Design of Air Ducts in Combustion Engines}.
    \item \textbf{***, 2021}: ROMSOC *** Workshop, Catania, Italy.
\end{itemize}

\section{Courses}
\begin{itemize}
    \item \textbf{Apr--Jun 2020}: Dr. \href{https://ifm.mathematik.uni-wuerzburg.de/~schmidt/}{Stephan Schmidt}. \textit{Computational Geometry and Shape Optimization}. Master course, Humboldt Universit\"at zu Berlin, 2020.
    \item \textbf{Oct 12--23, 2020}: Thematic Einstein Semester (TES): \href{https://mathplus.de/topic-development-lab/tes-winter-2020-21/}{Energy-based mathematical methods for reactive multiphase flows}.\newline $\triangleright$ TES Student Compact Course: \href{https://mathplus.de/topic-development-lab/tes-winter-2020-21/student-compact-course/}{Variational Methods for Fluids and Solids}.
\end{itemize}


\section{Talks}
\begin{itemize}
    \item \textbf{Sep 7, 2020}: Hong Nguyen (WIAS Berlin). ``A shape optimization problem for stationary Navier-Stokes flows in three-dimensional tubes'' [\href{https://zenodo.org/record/4117506}{slide}]. \href{https://morss2020.epfl.ch/}{Model Order Reduction Summer School 2020} (MORSS2020), \'Ecole polytechnique f\'ed\'erale de Lausanne (EPFL), Lausanne, Switzerland.
\end{itemize}

%------------------------------------------------------------------------------%

\printbibliography[heading=bibintoc]
\end{document}