\documentclass{article}
\usepackage[backend=biber,natbib=true,style=alphabetic,maxbibnames=50]{biblatex}
\addbibresource{/home/nqbh/reference/bib.bib}
\usepackage[utf8]{vietnam}
\usepackage{tocloft}
\renewcommand{\cftsecleader}{\cftdotfill{\cftdotsep}}
\usepackage[colorlinks=true,linkcolor=blue,urlcolor=red,citecolor=magenta]{hyperref}
\usepackage{amsmath,amssymb,amsthm,enumitem,float,graphicx,mathtools,tikz}
\usetikzlibrary{angles,calc,intersections,matrix,patterns,quotes,shadings}
\allowdisplaybreaks
\newtheorem{assumption}{Assumption}
\newtheorem{baitoan}{}
\newtheorem{cauhoi}{Câu hỏi}
\newtheorem{conjecture}{Conjecture}
\newtheorem{corollary}{Corollary}
\newtheorem{dangtoan}{Dạng toán}
\newtheorem{definition}{Definition}
\newtheorem{dinhly}{Định lý}
\newtheorem{dinhnghia}{Định nghĩa}
\newtheorem{example}{Example}
\newtheorem{ghichu}{Ghi chú}
\newtheorem{hequa}{Hệ quả}
\newtheorem{hypothesis}{Hypothesis}
\newtheorem{lemma}{Lemma}
\newtheorem{luuy}{Lưu ý}
\newtheorem{nhanxet}{Nhận xét}
\newtheorem{notation}{Notation}
\newtheorem{note}{Note}
\newtheorem{principle}{Principle}
\newtheorem{problem}{Problem}
\newtheorem{proposition}{Proposition}
\newtheorem{question}{Question}
\newtheorem{remark}{Remark}
\newtheorem{theorem}{Theorem}
\newtheorem{vidu}{Ví dụ}
\usepackage[left=1cm,right=1cm,top=5mm,bottom=5mm,footskip=4mm]{geometry}
\def\labelitemii{$\circ$}
\DeclareRobustCommand{\divby}{%
	\mathrel{\vbox{\baselineskip.65ex\lineskiplimit0pt\hbox{.}\hbox{.}\hbox{.}}}%
}
\setlist[itemize]{leftmargin=*}
\setlist[enumerate]{leftmargin=*}

\title{Philosophy $\Phi$ -- Triết Học $\Phi$}
\author{Nguyễn Quản Bá Hồng\footnote{A Scientist {\it\&} Creative Artist Wannabe. E-mail: {\tt nguyenquanbahong@gmail.com}. Bến Tre City, Việt Nam.}}
\date{\today}

\begin{document}
\maketitle
\begin{abstract}
	This text is a part of the series {\it Some Topics in Advanced STEM \& Beyond}:
	
	{\sc url}: \url{https://nqbh.github.io/advanced_STEM/}.
	
	Latest version:
	\begin{itemize}
		\item {\it Philosophy $\Phi$ -- Triết Học $\Phi$}.
		
		PDF: {\sc url}: \url{https://github.com/NQBH/advanced_STEM_beyond/blob/main/philosophy/NQBH_philosophy.pdf}.
		
		\TeX: {\sc url}: \url{https://github.com/NQBH/advanced_STEM_beyond/blob/main/philosophy/NQBH_philosophy.tex}.
	\end{itemize}
\end{abstract}
\tableofcontents

%------------------------------------------------------------------------------%

\section{Wikipedia}

\subsection{Wikipedia{\tt/}logical intuition}
``{\it Logical Intuition}, or {\it mathematical intuition} or {\it rational intuition}, is a series of instinctive foresight, know-how, \& savviness\footnote{{\it savvy} [a] having practical knowledge \& understanding of something; having common sense.} often associated with the ability to perceive \href{https://en.wikipedia.org/wiki/Logic}{logical} or mathematical truth -- \& the ability to solve mathematical challenges efficiency. Humans apply logical intuition in proving mathematical \href{https://en.wikipedia.org/wiki/Theorem}{theorems}, validating logical arguments, developing algorithms \& \href{https://en.wikipedia.org/wiki/Heuristic}{heuristics}, \& in related contexts where mathematical challenges are involved. The ability to recognize logical or mathematical truth \& identify viable methods may vary from person to person, \& may even be a result of knowledge \& experience, which are subject to cultivation. The ability may not be realizable in a computer program by means other than \href{https://en.wikipedia.org/wiki/Genetic_programming}{genetic programming} or \href{https://en.wikipedia.org/wiki/Evolutionary_programming}{evolutionary programming}.

\subsubsection{History}
\href{https://en.wikipedia.org/wiki/Plato}{Plato} \& \href{https://en.wikipedia.org/wiki/Aristotle}{Aristotle} considered intuition a means for perceiving ideas, significant enough that for Aristotle, intuition comprised the only means of knowing principles that are \href{https://en.wikipedia.org/wiki/A_priori_and_a_posteriori}{not subject to argument}.

\href{https://en.wikipedia.org/wiki/Henri_Poincar%C3%A9}{\sc Henri Poincar\'e} distinguished logical intuition from \href{https://en.wikipedia.org/wiki/Intuition}{other forms of intuition}. In his book \href{https://en.wikipedia.org/wiki/The_Value_of_Science}{The Value of Science}, he points out that:
\begin{quote}
	{\it``$\ldots$ There are many kinds of intuition. I have said how much the intuition of pure number, when comes rigorous mathematical induction, differs from sensible intuition to which the imagination, properly so called, is the {\it principal contributor}.''}
	
	-- $\ldots$ Có nhiều loại trực giác. Tôi đã nói rằng trực quan về số thuần túy, khi có quy nạp toán học nghiêm ngặt, khác biệt đến mức nào với trực giác nhạy cảm mà trí tưởng tượng, được gọi 1 cách chính xác, là {\it đóng góp chính}.	
\end{quote}
The passage goes on to assign 2 roles to logical intuition: to permit one to choose which \href{https://en.wikipedia.org/wiki/Axiom}{route} (axiom) to follow in search of scientific \href{https://en.wikipedia.org/wiki/Truth}{truth}, \& to allow one to comprehend logical developments.

\href{https://en.wikipedia.org/wiki/Bertrand_Russell}{\sc Bertrand Russell}, though critical of intuitive \href{https://en.wikipedia.org/wiki/Mysticism#Intuitive_insight_and_enlightenment}{mysticism}, pointed out that the degree to which a truth is \href{https://en.wikipedia.org/wiki/Self-evidence}{self-evident} according to logical intuition can vary, from 1 situation to another, \& stated that some self-evident truths are practically \href{https://en.wikipedia.org/wiki/Infallibility#In_philosophy}{infallible}\footnote{{\it infallible} [a] 1. never wrong; never making mistakes, $\ne$ {\it fallible}; 2. that never fails; always doing what it is supposed to do.}:
\begin{quote}
	{\it``When a certain number of logical principles have been admitted, the rest can be deduced from them; but the propositions deduced are often just as self-evident as those that were assumed without proof. All arithmetic, moreover, can be deduced from the general principles of logic, yet the simple propositions of arithmetic, such as `$2 + 2 = 4$', are just as self-evident as the principles of logic.''}
	
	-- Khi 1 số nguyên tắc logic nhất định đã được thừa nhận, phần còn lại có thể được suy ra từ chúng; nhưng các mệnh đề được suy ra thường hiển nhiên như những mệnh đề được giả định mà không có bằng chứng. Hơn nữa, tất cả số học đều có thể được suy ra từ các nguyên tắc chung của logic, tuy nhiên các mệnh đề đơn giản của số học, chẳng hạn như `$2 + 2 = 4$', cũng hiển nhiên như các nguyên tắc logic.
\end{quote}
\href{https://en.wikipedia.org/wiki/Kurt_G%C3%B6del}{\sc Kurt G\"odel} demonstrated based on his \href{https://en.wikipedia.org/wiki/G%C3%B6del%27s_incompleteness_theorems}{incompleteness theorems} that intuition-based \href{https://en.wikipedia.org/wiki/Propositional_calculus}{propositional calculus} cannot be \href{https://en.wikipedia.org/wiki/Many-valued_logic}{finitely valued}. G\"odel also likened logical intuition to sense perception, \& considered the mathematical constructs that humans perceive to have an independent \href{https://en.wikipedia.org/wiki/Philosophical_realism}{existence} of their own. Under this line of reasoning, the human mind's ability to sense such abstract constructs may not be finitely implementable.

\subsubsection{Discussion}
Dissent (bất đồng chính kiến) regarding the value of intuition in a logical or mathematical context may often hinge on the breadth of the definition of intuition \& the psychological underpinning of the word. Dissent regarding the implications of logical intuition in the fields of AI \& \href{https://en.wikipedia.org/wiki/Cognitive_computing}{cognitive computing} may similarly hinge on definitions. However, similarity between the potentially infinite nature of logical intuition posited by G\"odel \& the \href{https://en.wikipedia.org/wiki/Hard_problem_of_consciousness}{hard problem of consciousness} posited by \href{https://en.wikipedia.org/wiki/David_Chalmers}{\sc David Chalmers} suggest that the realms of intuitive knowledge \& experiential consciousness may both have aspects that are not reducible to classical physics concepts.'' -- \href{https://en.wikipedia.org/wiki/Logical_intuition}{Wikipedia{\tt/}logical intuition}

%------------------------------------------------------------------------------%

\subsection{Wikipedia{\tt/}abstract \& concrete}
``In philosophy \& the \href{https://en.wikipedia.org/wiki/The_arts}{arts},  a fundamental distinction is between things that are {\it abstract} \& things that are {\it concrete}. While there is no general \href{https://en.wikipedia.org/wiki/Consensus_decision-making}{consensus} as to how to precisely defined the 2, examples include that things like \href{https://en.wikipedia.org/wiki/Number}{numbers}, \href{https://en.wikipedia.org/wiki/Set_(mathematics)}{sets}, \& \href{https://en.wikipedia.org/wiki/Idea}{ideas} are abstract objects, while plants, dogs, \& planets are concrete objects. Popular suggestions for a definition include that the distinction between concreteness vs. abstractness is, respectively: between
\begin{enumerate}
	\item existence inside vs. outside \href{https://en.wikipedia.org/wiki/Spacetime}{space-time};
	\item having causes \& effects vs. not;
	\item being related, in \href{https://en.wikipedia.org/wiki/Metaphysics}{metaphysics}, to \href{https://en.wikipedia.org/wiki/Particulars}{particulars} vs. \href{https://en.wikipedia.org/wiki/Universal_(metaphysics)}{universals};
	\item belonging to either the physical vs. the mental realm (or the mental-\&-physical realm vs. neither).
\end{enumerate}
Another view is that it is the distinction between contingent existence vs. necessary existence; however, philosophers differ on which type of existence here defines abstractness, as opposed to concreteness. Despite this diversity of views, there is broad agreement concerning most objects as to whether they are abstract or concrete, s.t. most interpretations agree, e.g., that rocks are concrete objects while numbers are abstract objects.

Abstract objects are most commonly used in philosophy, particularly metaphysics, \& \href{https://en.wikipedia.org/wiki/Semantics}{semantics}. They are sometimes called {\it abstracta} in contrast to {\it concreta}. The term {\it abstract object} is said to have been coined by \href{https://en.wikipedia.org/wiki/Willard_Van_Orman_Quine}{\sc Willard Van Orman Quine}. \href{https://en.wikipedia.org/wiki/Abstract_object_theory}{Abstract object theory} is a discipline that studies the nature \& role of abstract objects. It holds that properties can be related to objects in 2 ways: through exemplification \& through encoding. Concrete objects exemplify their properties while abstract objects merely encode them. This approach is also known as the \href{https://en.wikipedia.org/wiki/Dual_copula_strategy}{dual copula strategy}.

\subsubsection{In philosophy}
The \href{https://en.wikipedia.org/wiki/Type%E2%80%93token_distinction}{type-token distinction} identifies physical objects that are tokens of a particular type of thing. The ``type'' of which it is a part is in itself an abstract object. The abstract-concrete distinction is often introduced \& initially understood in terms of \href{https://en.wikipedia.org/wiki/Paradigm}{paradigmatic} examples of objects of each kind. Examples of abstract vs. concrete objects: tennis vs. a tennis match, redness vs. red light reflected off of an apple \& hitting one's eyes, 5 vs. 5 cars, justice vs. a just action, humanity (the property of being human) vs. human population (the set of all humans).

Abstract objects have often garnered the interest of philosophers because they raise problems for popular theories. In \href{https://en.wikipedia.org/wiki/Ontology}{ontology}, abstract objects are considered problematic for \href{https://en.wikipedia.org/wiki/Physicalism}{physicalism} \& some forms of \href{https://en.wikipedia.org/wiki/Metaphysical_naturalism}{naturalism}. Historically, the most important ontological dispute about abstract objects has been the \href{https://en.wikipedia.org/wiki/Problem_of_universals}{problem of universals}. In \href{https://en.wikipedia.org/wiki/Epistemology}{epistemology}, abstract objects are considered problematic for \href{https://en.wikipedia.org/wiki/Empiricism}{empiricism}. If abstracta lack causal powers \& spatial location, how do we know about them? It is hard to say how they can affect our sensory experiences, \& yet we seem to agree on a wide range of claims about them.

Some, e.g. \href{https://en.wikipedia.org/wiki/Ernst_Mally}{\sc Ernst Mally}, \href{https://en.wikipedia.org/wiki/Edward_Zalta}{\sc Edward Zalta}, \& arguably, \href{https://en.wikipedia.org/wiki/Plato}{\sc Plato} in his \href{https://en.wikipedia.org/wiki/Theory_of_Forms}{\it Theory of Forms}, have held that abstract objects constitute the defining subject matter of \href{https://en.wikipedia.org/wiki/Metaphysics}{metaphysics} or philosophical inquiry more broadly. To the extent that philosophy is independent of empirical research, \& to the extent that empirical questions do not inform questions about abstracta, philosophy would seem especially suited to answering these latter questions.

In \href{https://en.wikipedia.org/wiki/Modern_philosophy}{modern philosophy}, the distinction between abstract \& concrete was explored by \href{https://en.wikipedia.org/wiki/Immanuel_Kant}{\sc Immanuel Kant} \& \href{https://en.wikipedia.org/wiki/G._W._F._Hegel}{G. W. F. Hegel}.

\href{https://en.wikipedia.org/wiki/Gottlob_Frege}{\sc Gottlob Frege} said that abstract objects, e.g., propositions, were members of $\frac{1}{3}$ realm, different from the external world or from internal \href{https://en.wikipedia.org/wiki/Consciousness}{consciousness}. See \href{https://en.wikipedia.org/wiki/Popper%27s_three_worlds}{{\sc Popper}'s 3 worlds}.

\paragraph{Abstract objects \& causality.} Another popular proposal for drawing the abstract-concrete distinction contends that an object is abstract if it lacks \href{https://en.wikipedia.org/wiki/Causality}{causal} power. A causal power has the ability to affect something causally. Thus, the empty set is abstract because it cannot act on other objects. 1 problem with this view is that it is not clear exactly what it is to have causal power. For a more detailed exploration of the abstract-concrete distinction, see the relevant \href{https://en.wikipedia.org/wiki/Stanford_Encyclopedia_of_Philosophy}{\it Stanford Encyclopedia of Philosophy} article.

\paragraph{Quasi-abstract entities.} Recently, there has been some philosophical interest in the development of a 3rd category of objects known as the quasi-abstract. Quasi-abstract objects have drawn particular attention in the area of \href{https://en.wikipedia.org/wiki/Social_ontology}{social ontology} \& \href{https://en.wikipedia.org/wiki/Documentality}{documentality}. Some argue that the over-adherence to the \href{https://en.wikipedia.org/wiki/Platonism}{platonist} duality of the concrete \& the abstract has led to a large category of social objects having been overlooked or rejected as \href{https://en.wikipedia.org/wiki/Nonexistent_object}{nonexistent} because they exhibit characteristics that the traditional duality between concrete \& abstract regards as incompatible. Specifically, the ability to have temporal location, but not spatial location, \& have causal agency (if only by acting through representatives). These characteristics are exhibited by a number of social objects, including states of the international legal system.

\subsubsection{Concrete \& abstract thought in psychology}
\href{https://en.wikipedia.org/wiki/Jean_Piaget}{\sc Jean Piaget} uses the terms ``concrete'' \& ``formal'' to describe 2 different types of learning. Concrete thinking involves facts \& descriptions about everyday, tangible objects, while abstract (\href{https://en.wikipedia.org/wiki/Formal_Operational#Formal_operational_stage}{formal operational}) thinking involves a mental process. Abstract idea vs. Concrete idea: dense things sink vs. It will sink if its density is greater than the density of the fluid, you breathe in oxygen \& breathe out carbon dioxide vs. gas exchange takes place between the air in the alveoli \& the blood, plants get water through their roots vs. water diffuses through the cell membrane of the root hair cells.'' -- \href{https://en.wikipedia.org/wiki/Abstract_and_concrete}{Wikipedia{\tt/}abstract \& concrete}

%------------------------------------------------------------------------------%

\subsection{Wikipedia{\tt/}natural law}
``{\it Natural law} (Latin: {\it ius naturale, lex naturalis}) is a system of law based on a close observation of \href{https://en.wikipedia.org/wiki/Natural_order_(philosophy)}{natural order} \& \href{https://en.wikipedia.org/wiki/Human_nature}{human nature}, from which values, thought by natural law's proponents to be intrinsic to human nature, can be deduced \& applied independently of \href{https://en.wikipedia.org/wiki/Positive_law}{positive law} (the express enacted laws of a \href{https://en.wikipedia.org/wiki/Sovereign_state}{state} or \href{https://en.wikipedia.org/wiki/Society}{society}). According to the theory of law called \href{https://en.wikipedia.org/wiki/Jusnaturalism}{jusnaturalism}, all people have inherent rights, conferred not by act of legislation but by ``\href{https://en.wikipedia.org/wiki/God}{God}, \href{https://en.wikipedia.org/wiki/Nature}{nature}, or \href{https://en.wikipedia.org/wiki/Reason}{reason}''. Natural law theory can also refer to ``theories of ethics, theories of politics, theories of civil law, \& theories of religious morality.''
	
\subsubsection{History}

\subsubsection{Contemporary jurisprudence}

\subsubsection{Methodology}

'' -- \href{https://en.wikipedia.org/wiki/Natural_law}{Wikipedia{\tt/}natural law}

%------------------------------------------------------------------------------%

\subsection{Wikipedia{\tt/}universal law}
``In \href{https://en.wikipedia.org/wiki/Law}{law} \& \href{https://en.wikipedia.org/wiki/Ethics}{ethics}, {\it universal law} or {\it universal principle} refers to concepts of legal \href{https://en.wikipedia.org/wiki/Legitimacy_(political)}{legitimacy} actions, whereby those \href{https://en.wikipedia.org/wiki/Value_(personal_and_cultural)}{principles} \& rules for governing human beings' conduct which are most universal in their acceptability, their applicability, translation, \& \href{https://en.wikipedia.org/wiki/Philosophical}{philosophical} basis, are therefore considered to be most legitimate.

\subsubsection{Debate}
Cognition, experiences \& intuition are the starting points of legal thought, which has to be seen through the glasses of universality \& abstractness. Notwithstanding this assumption, ``legal principles
\begin{enumerate}
	\item do not contain only logic \& reason \&
	\item they can be different in different situations despite their equal naming.
\end{enumerate}
The legal rules can be identical in different legal orders while they carry different wants''.

On 1 side ``universality, abstraction, \& theory itself are defined in a way that undermines the perspectives of some while privileging the perspectives of others''; on the other side, ``the aspiration to universality itself may stand in the way of its realization if it seals off from view the bias built into legal norms, public practices, \& established institutions''.

\subsubsection{Examples}
There are 12 universal laws.
\begin{enumerate}
	\item Law of Divine Oneness
	\item Law of Vibration
	\item Law of Action
	\item Law of Correspondence
	\item Law of Cause \& Effect
	\item Law of Compensation
	\item Law of Attraction
	\item Law of Perpetual Transmutation of Energy
	\item Law of Relativity
	\item Law of Polarity
	\item Law of Rhythm
	\item Law of Gender
\end{enumerate}
'' -- \href{https://en.wikipedia.org/wiki/Universal_law}{Wikipedia{\tt/}universal law}

%------------------------------------------------------------------------------%

\subsection{Wikipedia{\tt/}universality (philosophy)}
``In philosophy, {\it universality} or {\it absolutism} is the idea that universal facts exist \& can be progressively discovered, as opposed to \href{https://en.wikipedia.org/wiki/Relativism}{relativism}, which asserts that all facts are relative to one's perspective. Absolutism \& relativism have been explored at length in contemporary \href{https://en.wikipedia.org/wiki/Analytic_philosophy}{analytic philosophy}.

\subsubsection{Universality in ethics}
Main article: \href{https://en.wikipedia.org/wiki/Moral_universalism}{Wikipedia{\tt/}moral universalism}. When used in the context of ethics, the meaning of {\it universal} refers to that which is true for ``all similarly situated individuals''. \href{https://en.wikipedia.org/wiki/Rights}{Rights}, e.g. in \href{https://en.wikipedia.org/wiki/Natural_rights}{natural rights}, or in the 1789 \href{https://en.wikipedia.org/wiki/Declaration_of_the_Rights_of_Man_and_of_the_Citizen}{Declaration of the Rights of Man \& of the Citizen}, for those heavily influenced by the philosophy of the \href{https://en.wikipedia.org/wiki/Age_of_Enlightenment}{Enlightenment} \& its conception of a \href{https://en.wikipedia.org/wiki/Human_nature}{human nature}, could be considered universal. The 1948 \href{https://en.wikipedia.org/wiki/Universal_Declaration_of_Human_Rights}{Universal Declaration of Human Rights} is inspired by such principles.

Universal moralities contrast with \href{https://en.wikipedia.org/wiki/Moral_relativism}{moral relativisms}, which seek to account for differing ethical positions between people \& \href{https://en.wikipedia.org/wiki/Social_norm}{cultural norms}.

\subsubsection{Universality about truth}
In logic, or the consideration of valid arguments, a \href{https://en.wikipedia.org/wiki/Proposition}{proposition} is said to have universality if it can be conceived as being true in all possible contexts without creating a \href{https://en.wikipedia.org/wiki/Contradiction}{contradiction}. A \href{https://en.wikipedia.org/wiki/Universalism}{universalist conception of truth} accepts 1 or more universals, whereas a \href{https://en.wikipedia.org/wiki/Relativism}{relativist conception of truth} denies the existence of some or all \href{https://en.wikipedia.org/wiki/Problem_of_universals}{universals}.

\subsubsection{Universal in metaphysics}
Main article: \href{https://en.wikipedia.org/wiki/Universal_(metaphysics)}{Wikipedia{\tt/}universal (metaphysics)}. In \href{https://en.wikipedia.org/wiki/Metaphysics}{metaphysics}, a \href{https://en.wikipedia.org/wiki/Universal_(metaphysics)}{universal} is a proposed \href{https://en.wikipedia.org/wiki/Type_(metaphysics)}{type}, \href{https://en.wikipedia.org/wiki/Property_(metaphysics)}{property}, or \href{https://en.wikipedia.org/wiki/Relation_(metaphysics)}{relation} which can be instantiated by many different \href{https://en.wikipedia.org/wiki/Particular}{particulars}. While universals are related to the concept o universality, the concept is importantly distinct, see the main page on universals for a full treatment of the topic.'' -- \href{https://en.wikipedia.org/wiki/Universality_(philosophy)}{Wikipedia{\tt/}universality (philosophy)}

%------------------------------------------------------------------------------%

\section{{\sc Blaise Pascal}. {\it Pens\'ees} \& Other Writings}

\begin{enumerate}
	\item \cite{Pascal_pensees}.{\sc Blaise Pascal}. {\it Pens\'ees}.
	
	{\sf Note.} {\it Passages} erased by {\sc Pascal} are enclosed in square brackets, thus []. {\it Words}, added or corrected by the editor of the text, are similarly denoted, but are in italics. It has been seen fit to transfer Fragment 514 of the French edition to the Notes. All subsequent Fragments have accordingly been renumbered.
	\begin{itemize}
		\item {\sf Introduction by {\sc T. S. Eliot}.} It might seem that about {\sc Blaise Pascal}, \& about the 2 works on which his fame is founded, everything that there is to say had been said. The details of his life are as fully known as we can expect to know them; his mathematical \& physical discoveries have been treated many times; his religious sentiment \& his theological views have been discussed again \& again; \& his prose style has been analyzed by French critics down to the finest particular. But {\sc Pascal} is 1 of those writers who will be \& who must be studied afresh by men in every generation. It is not he who changes, but we who change. It is not our knowledge of him that increases, but our world that alters \& our attitudes towards it. The history of human opinions of {\sc Pascal} \& of men of his stature is a part of the history of humanity. That indicates his permanent importance.
		
		-- Có vẻ như về {\sc Blaise Pascal}, \& về 2 tác phẩm mà danh tiếng của ông được xây dựng trên đó, mọi thứ cần nói đều đã được nói đến. Chi tiết về cuộc đời của ông được biết đến đầy đủ như chúng ta có thể mong đợi biết; những khám phá toán học \& vật lý của ông đã được đề cập nhiều lần; tình cảm tôn giáo \& quan điểm thần học của ông đã được thảo luận lại \& nhiều lần; \& phong cách văn xuôi của ông đã được các nhà phê bình người Pháp phân tích đến từng chi tiết nhỏ nhất. Nhưng {\sc Pascal} là 1 trong những nhà văn sẽ \& phải được những người đàn ông trong mỗi thế hệ nghiên cứu lại. Không phải ông ấy thay đổi, mà là chúng ta thay đổi. Không phải kiến thức của chúng ta về ông ấy tăng lên, mà thế giới của chúng ta thay đổi \& thái độ của chúng ta đối với ông ấy. Lịch sử về quan điểm của con người về {\sc Pascal} \& những người có tầm vóc như ông là một phần của lịch sử nhân loại. Điều đó cho thấy tầm quan trọng lâu dài của ông.
		
		The facts of {\sc Pascal}'s life, so far as they are necessary for this brief introduction to the {\it Pens\'ees}, are as follows. He was born at Clermont, in Auvergne, in 1623. His family were people of substance of the upper middle class. His father was a government official, who was able to leave, when he died, a sufficient patrimony to his 1 son \& his 2 daughters. In 1631 the feather moved to Paris, \& a few years later took up another government post at Rouen. Wherever he lived, the elder {\sc Pascal} seems to have mingled with some of the best society, \& with men of eminence in science \& the arts. {\sc Blaise} was educated entirely by his father at home. He was exceedingly precocious, indeed excessively precocious, for his application to studies in childhood \& adolescence impaired his health, \& is held responsible for his death at 39. Prodigious, thought not incredible stories are preserved, especially of his precocity in mathematics. His mind was active rather than accumulative; he showed from his earliest years that disposition to find things out for himself, which has characterized the infancy of {\sc Clerk-Maxwell} \& other scientists. Of his later discoveries in physics there is no need for mention here; it must only be remembered that he counts as 1 of the greatest physicists \& mathematicians of all time; \& that his discoveries were made during the years when most scientists are still apprentices.
		
		-- Những sự kiện trong cuộc đời của {\sc Pascal}, trong chừng mực cần thiết cho phần giới thiệu ngắn gọn này về {\it Pens\'ees}, như sau. Ông sinh ra tại Clermont, ở Auvergne, vào năm 1623. Gia đình ông là những người có thế lực thuộc tầng lớp trung lưu thượng lưu. Cha ông là một viên chức chính phủ, người đã để lại, khi ông qua đời, một gia tài đủ lớn cho 1 người con trai \& 2 người con gái của ông. Năm 1631, ông chuyển đến Paris, \& vài năm sau đó đảm nhiệm một chức vụ chính phủ khác tại Rouen. Bất cứ nơi nào ông sống, {\sc Pascal} lớn tuổi hơn dường như đã hòa nhập với một số người trong xã hội tốt nhất, \& với những người đàn ông lỗi lạc trong khoa học \& nghệ thuật. {\sc Blaise} được cha mình giáo dục hoàn toàn tại nhà. Ông cực kỳ thông minh sớm, thực sự là quá thông minh sớm, vì việc ông áp dụng vào việc học trong thời thơ ấu \& thời niên thiếu đã làm suy yếu sức khỏe của ông, \& bị cho là nguyên nhân gây ra cái chết của ông ở tuổi 39. Những câu chuyện phi thường, mặc dù không phải là khó tin, vẫn được lưu giữ, đặc biệt là về sự thông minh sớm của ông trong toán học. Trí óc của ông năng động hơn là tích lũy; ông đã thể hiện từ những năm đầu đời rằng mình có khuynh hướng tự mình tìm hiểu mọi thứ, điều này đã đặc trưng cho thời thơ ấu của {\sc Clerk-Maxwell} \& các nhà khoa học khác. Về những khám phá sau này của ông trong vật lý, không cần phải đề cập ở đây; chỉ cần nhớ rằng ông được coi là 1 trong những nhà vật lý \& toán học vĩ đại nhất mọi thời đại; \& rằng những khám phá của ông được thực hiện trong những năm mà hầu hết các nhà khoa học vẫn còn là người học việc.
		
		The elder {\sc Pascal, \'Etienne}, was a sincere Christian. About 1646 he fell in with some representatives of the religious revival within the Church which has become known as Jansenism -- after Jansenius, Bishop of Ypres, whose theological work is taken as the origin of the movement. This period is usually spoken of as the moment of {\sc Pascal}'s ``1st conversion.'' The word ``conversion,'' however, is too forcible to be applied at this point to {\sc Blaise Pascal} himself. The family had always been devout, \& the young {\sc Pascal}, though absorbed in his scientific work, never seems to have been afflicted with infidelity. His attention was then directed, certainly, to religious \& theological matters; but the term ``conversion'' can only be applied to his sisters -- the elder, already Madame P\'erier, \& particularly the younger, {\sc Jacqueline}, who at that time conceived a vocation for the religious life. {\sc Pascal} himself was by no means disposed to renounce the world. After the death of the father in 1650 {\sc Jacqueline}, a young woman of remarkable strength \& beauty of character, wished to take her vows as a sister of Port-Royal, \& for some time her wish remained unfulfilled owing to the opposition of her brother. His objection was on the purely worldly ground that she wished to make over her patrimony to the Order; whereas while she lived with him, their combined sources made it possible for him to live more nearly on a scale of expense congenial to his tastes. He liked, in fact, not only to mix with the best society, but to keep a coach \& horses -- 6 horses is the number at 1 time attributed to his carriage. Though he had no legal power to prevent his sister from disposing of her property as she elected, the amiable {\sc Jacqueline} shrank from doing so without her brother's willing approval. The Mother Superior, M\`ere Ang\'elique -- herself an eminent personage in the history of this religious movement -- finally persuaded the young novice to enter the order without the satisfaction of bringing her patrimony with her; but {\sc Jacqueline} remained so distressed by this situation that her brother finally relented.
		
		-- Người anh cả {\sc Pascal, \'Etienne}, là một Cơ đốc nhân chân thành. Khoảng năm 1646, ông đã gặp một số đại diện của phong trào phục hưng tôn giáo trong Giáo hội, được gọi là Jansenism -- theo tên của Jansenius, Giám mục Ypres, người có tác phẩm thần học được coi là nguồn gốc của phong trào này. Giai đoạn này thường được coi là thời điểm ``lần cải đạo đầu tiên'' của {\sc Pascal}. Tuy nhiên, từ ``cải đạo'' quá mạnh mẽ để có thể áp dụng vào thời điểm này cho chính {\sc Blaise Pascal}. Gia đình luôn sùng đạo, \& chàng trai trẻ {\sc Pascal}, mặc dù đắm chìm trong công trình khoa học của mình, nhưng dường như chưa bao giờ bị ảnh hưởng bởi sự bất trung. Chắc chắn là khi đó, sự chú ý của ông hướng đến các vấn đề tôn giáo \& thần học; nhưng thuật ngữ ``cải đạo'' chỉ có thể áp dụng cho các chị gái của ông -- người chị, đã là Madame P\'erier, \& đặc biệt là người em, {\sc Jacqueline}, người vào thời điểm đó đã hình thành ơn gọi cho cuộc sống tu trì. Bản thân {\sc Pascal} không hề có ý định từ bỏ thế gian. Sau cái chết của người cha vào năm 1650 {\sc Jacqueline}, một phụ nữ trẻ có sức mạnh đáng kinh ngạc \& vẻ đẹp của tính cách, muốn tuyên thệ trở thành nữ tu của Port-Royal, \& trong một thời gian, mong muốn của cô vẫn chưa được thực hiện do sự phản đối của anh trai cô. Sự phản đối của anh ta là trên cơ sở hoàn toàn thế tục rằng cô muốn chuyển giao tài sản của mình cho Dòng; trong khi khi cô sống với anh ta, nguồn thu nhập chung của họ giúp anh ta có thể sống gần hơn với mức chi tiêu phù hợp với sở thích của mình. Trên thực tế, anh ta không chỉ thích giao du với những người thượng lưu nhất mà còn thích có một cỗ xe ngựa \& ngựa -- 6 con ngựa là số lượng tại 1 thời điểm được cho là dành cho cỗ xe của anh ta. Mặc dù không có thẩm quyền pháp lý để ngăn cản chị gái mình định đoạt tài sản của bà theo ý bà, {\sc Jacqueline} đáng mến đã không làm như vậy nếu không có sự chấp thuận của anh trai bà. Mẹ Bề trên, M\`ere Ang\'elique -- bản thân bà là một nhân vật nổi tiếng trong lịch sử của phong trào tôn giáo này -- cuối cùng đã thuyết phục được cô tập sự trẻ tuổi gia nhập dòng mà không hài lòng khi mang theo gia tài của mình; nhưng {\sc Jacqueline} vẫn rất đau khổ vì tình huống này đến nỗi cuối cùng anh trai bà đã nhượng bộ.
		
		So far as is known, the worldly life enjoyed by {\sc Pascal} during this period can hardly be qualified as ``dissipation,'' \& certainly not as ``debauchery.'' Even gambling may have appealed to him chiefly as affording a study of mathematical probabilities. He appeared to have led such a life as any cultivated intellectual man of good position \& independent means might lead \& consider himself a model of probity \& virtue. Not even a love-affair is laid at his door, though he is said to have contemplated marriage. But Jansenism, as represented by the religious society of Port-Royal, was morally a Puritan movement within the Church, \& its standards of conduct were at least as severe as those of any Puritanism in England or America. The period of fashionable society, in {\sc Pascal}'s life, is however, of great importance in his development. It enlarged his knowledge of men \& refined his tastes; he became a man of the world \& never lost what he had learnt; \& when he turned his thoughts wholly towards religion, his worldly knowledge was a part of his composition which is essential to the value of his work.
		
		-- Theo như được biết, cuộc sống thế tục mà {\sc Pascal} tận hưởng trong thời kỳ này khó có thể được coi là ``phung phí'', \& chắc chắn không phải là ``trụy lạc''. Ngay cả cờ bạc cũng có thể hấp dẫn ông chủ yếu vì giúp ông nghiên cứu về xác suất toán học. Ông dường như đã sống một cuộc sống mà bất kỳ người đàn ông trí thức có học thức nào có địa vị tốt \& phương tiện độc lập có thể sống \& coi mình là hình mẫu của sự liêm chính \& đức hạnh. Thậm chí không có một mối tình nào được đặt trước cửa ông, mặc dù ông được cho là đã cân nhắc đến hôn nhân. Nhưng Jansenism, được đại diện bởi xã hội tôn giáo Port-Royal, về mặt đạo đức là một phong trào Thanh giáo trong Giáo hội, \& các tiêu chuẩn ứng xử của nó ít nhất cũng nghiêm ngặt như bất kỳ Thanh giáo nào ở Anh hoặc Mỹ. Tuy nhiên, giai đoạn xã hội thời thượng, trong cuộc đời của {\sc Pascal}, có tầm quan trọng lớn trong sự phát triển của ông. Nó mở rộng kiến thức của ông về con người \& tinh chỉnh thị hiếu của ông; ông trở thành một người của thế giới \& không bao giờ đánh mất những gì mình đã học được; \& khi ông hướng toàn bộ suy nghĩ của mình tới tôn giáo, kiến thức thế gian của ông là một phần trong tác phẩm của ông, điều này rất cần thiết cho giá trị của tác phẩm.
		
		{\sc Pascal}'s interest in society did not distract him from scientific research; nor did this period occupy much space in what is a very short \& crowded life. Partly his natural dissatisfaction with such a life, once he had learned all it had to teach him, partly the influence of his saintly sister {\sc Jacqueline}, partly increasing suffering as his health declined, directed him more \& more out of the world \& to thoughts of eternity. \& in 1654 occurs what is called his ``2nd conversion,'' but which might be called his conversion simply.
		
		-- Sự quan tâm của {\sc Pascal} đối với xã hội không làm ông sao nhãng khỏi nghiên cứu khoa học; giai đoạn này cũng không chiếm nhiều không gian trong cuộc đời rất ngắn ngủi \& đông đúc. Một phần là do sự bất mãn tự nhiên của ông với cuộc sống như vậy, sau khi ông đã học được tất cả những gì nó phải dạy cho ông, một phần là do ảnh hưởng của người chị thánh thiện {\sc Jacqueline}, một phần là do sự đau khổ gia tăng khi sức khỏe của ông suy yếu, đã hướng ông \& nhiều hơn ra khỏi thế giới \& hướng đến những suy nghĩ về cõi vĩnh hằng. \& vào năm 1654, điều được gọi là ``sự cải đạo lần thứ 2'' của ông, nhưng có thể gọi đơn giản là sự cải đạo của ông.
		
		He made a note of his mystical experience, which he kept always about him, \& which was found, after his death, sewn into the coat which he was wearing. The experience occurred on Nov 23, 1654, \& there is no reason to doubt its genuineness unless we choose to deny all mystical experience. Now, {\sc Pascal} was not a mystic, \& his works are not to be classified amongst mystical writings; but what can only be called mystical experience happens to many men who do not become mystics. The work which he undertook soon after, the {\it Lettres \'ecrites \`a un provincial}, is a masterpiece of religious controversy at the opposite pole from mysticism. We know quite well that he was at the time when he received his illumination from God in extremely poor health; but it is a commonplace that some forms of illness are extremely favorable, not only to religious illumination, but to artistic \& literary composition. A piece of writing meditated, apparently without progress, for months or years, may suddenly take shape \& word; \& in this state long passages may be produced which require little or no retouch. I have no good word to say for the cultivation of automatic writing as the model of literary composition; I doubt whether these moments {\it can} be cultivated by the writer; but he to whom this happens assuredly has the sensation of being a vehicle rather than a maker. No masterpiece can be produced whole by such means; but neither does even the higher form of religious inspiration suffice for the religious life; even the most exalted mystic must return to the world, \& use his reason to employ the results of his experience in daily life. You may call it communion with the Divine, or you may call it a temporary crystallization of the mind. Until science can teach us to reproduce such phenomena at will, science cannot claim to have explained them; \& they can be judged only by their fruits.
		
		-- Ông đã ghi chép lại trải nghiệm thần bí của mình, mà ông luôn giữ bên mình, \& sau khi ông qua đời, được tìm thấy, được khâu vào chiếc áo khoác mà ông đang mặc. Trải nghiệm này xảy ra vào ngày 23 tháng 11 năm 1654, \& không có lý do gì để nghi ngờ tính xác thực của nó trừ khi chúng ta chọn cách phủ nhận mọi trải nghiệm thần bí. Bây giờ, {\sc Pascal} không phải là một nhà thần bí, \& các tác phẩm của ông không được phân loại vào các tác phẩm thần bí; nhưng những gì chỉ có thể được gọi là trải nghiệm thần bí xảy ra với nhiều người đàn ông không trở thành nhà thần bí. Tác phẩm mà ông thực hiện ngay sau đó, {\it Lettres \'ecrites \`a un province}, là một kiệt tác của cuộc tranh luận tôn giáo ở cực đối lập với chủ nghĩa thần bí. Chúng ta biết khá rõ rằng vào thời điểm ông nhận được sự khai sáng từ Chúa, sức khỏe của ông cực kỳ yếu; nhưng điều thường thấy là một số dạng bệnh tật lại cực kỳ có lợi, không chỉ đối với sự khai sáng tôn giáo, mà còn đối với sáng tác nghệ thuật \& văn học. Một tác phẩm được suy ngẫm, dường như không có tiến triển, trong nhiều tháng hoặc nhiều năm, có thể đột nhiên thành hình \& từ; \& trong trạng thái này, những đoạn văn dài có thể được tạo ra mà ít hoặc không cần chỉnh sửa. Tôi không có lời nào hay để nói về việc bồi dưỡng lối viết tự động như mô hình sáng tác văn học; Tôi nghi ngờ liệu những khoảnh khắc này {\it có thể} được người viết bồi dưỡng hay không; nhưng người mà điều này xảy ra chắc chắn có cảm giác như mình là một phương tiện hơn là một người sáng tạo. Không có kiệt tác nào có thể được tạo ra trọn vẹn bằng những phương tiện như vậy; nhưng ngay cả hình thức cảm hứng tôn giáo cao hơn cũng không đủ cho cuộc sống tôn giáo; ngay cả nhà huyền môn cao quý nhất cũng phải trở về với thế giới, \& sử dụng lý trí của mình để sử dụng kết quả kinh nghiệm của mình trong cuộc sống hàng ngày. Bạn có thể gọi đó là sự giao tiếp với Đấng thiêng liêng, hoặc bạn có thể gọi đó là sự kết tinh tạm thời của tâm trí. Cho đến khi khoa học có thể dạy chúng ta tái tạo những hiện tượng như vậy theo ý muốn, khoa học không thể tuyên bố đã giải thích được chúng; \& chúng chỉ có thể được đánh giá bằng thành quả của chúng.
		
		From that time until his death, {\sc Pascal} was closely associated with the society of Port-Royal which his sister {\sc Jacqueline}, who predeceased him, had joined as a {\it religieuse}; the society was then fighting for its life against the Jesuits. 5 propositions, judged by a committee of cardinals \& theologians at Rome to be heretical, were found to be put forward in the work of Jansenius; \& the society of Port-Royal, the representative of Jansenism among devotional communities, suffered a blow from which it never revived. It is not the place here to review the bitter controversy \& conflict; the best account, from the point of view of a critic of genius who took no side, who was neither Jansenist nor Jesuit, Christian nor infidel, is that in the great book of {\sc Sainte-Beuve}, {\it Port-Royal}. \& in this book the parts devoted to {\sc Pascal} himself are among the most brilliant pages of criticism that {\sc Sainte-Beuve} ever wrote. It is sufficient to notice that the next occupation of {\sc Pascal}, after his conversion, was to write these 18 ``Letters,'' which as prose are of capital importance in the foundation of French classical style, \& which as polemic are surpassed by none, not by {\sc Demonsthenes}, or {\sc Cicero}, or {\sc Swift}. They have the limitation of all polemic \& forensic: they persuade, they seduce, they are unfair. But it is also unfair to assert that, in these {\it Letters to a Provincial}, {\sc Pascal} was attacking the Society of Jesus in itself. He was attacking rather a particular school of casuistry which relaxed the requirements of the Confessional; a school which certainly flourished amongst the Society of Jesus at that time, \& of which the {\sc Spaniards Escobar \& Molina} are the most eminent authorities. He undoubtedly abused the art of quotation, as a polemical writer can hardly help but do; but there were abuses for him to abuse; \& he did the job thoroughly. His {\it Letters} must not be called theology. Academic theology was not a department in which {\sc Pascal} was versed; when necessary, the fathers of Port-Royal came to his aid. The {\it Letters} are the work of 1 of the finest mathematical minds of any time, \& of a man of the world who addressed, not theologians, but the world in general -- all of the cultivated \& many of the less cultivated of the French laity; \& with this public they made an astonishing success.
		
		-- Từ thời điểm đó cho đến khi qua đời, {\sc Pascal} gắn bó chặt chẽ với hội Port-Royal mà chị gái ông {\sc Jacqueline}, người đã mất trước ông, đã gia nhập với tư cách là một {\it religieuse}; hội này khi đó đang đấu tranh để sinh tồn chống lại các tu sĩ Dòng Tên. 5 đề xuất, được một ủy ban gồm các hồng y \& các nhà thần học tại Rome phán quyết là dị giáo, đã được đưa ra trong tác phẩm của Jansenius; \& hội Port-Royal, đại diện của Jansenism trong các cộng đồng sùng đạo, đã phải chịu một đòn mà nó không bao giờ hồi sinh. Đây không phải là nơi để xem xét lại cuộc tranh cãi \& xung đột cay đắng; lời tường thuật hay nhất, theo quan điểm của một nhà phê bình thiên tài không đứng về phe nào, không theo Jansenist cũng không theo Jesuit, không theo Cơ đốc giáo cũng không theo đạo ngoại, là trong cuốn sách tuyệt vời của {\sc Sainte-Beuve}, {\it Port-Royal}. \& trong cuốn sách này, những phần dành riêng cho {\sc Pascal} nằm trong số những trang phê bình sáng chói nhất mà {\sc Sainte-Beuve} từng viết. Chỉ cần lưu ý rằng công việc tiếp theo của {\sc Pascal}, sau khi cải đạo, là viết 18 ``Bức thư'' này, với tư cách là văn xuôi, có tầm quan trọng hàng đầu trong nền tảng của phong cách cổ điển Pháp, \& là những tác phẩm có tính tranh luận không ai vượt qua được, kể cả {\sc Demonsthenes}, hay {\sc Cicero}, hay {\sc Swift}. Chúng có hạn chế của mọi tác phẩm tranh luận \& pháp y: chúng thuyết phục, chúng quyến rũ, chúng bất công. Nhưng cũng không công bằng khi khẳng định rằng, trong {\it Letters to a Provincial} này, {\sc Pascal} đang tấn công chính Dòng Tên. Ông đang tấn công một trường phái ngụy biện cụ thể vốn đã nới lỏng các yêu cầu của Hội đồng Tuyên tín; một trường phái chắc chắn đã phát triển mạnh mẽ trong Hội Jesus vào thời điểm đó, \& trong đó {\sc Người Tây Ban Nha Escobar \& Molina} là những người có thẩm quyền nổi tiếng nhất. Ông chắc chắn đã lạm dụng nghệ thuật trích dẫn, như một nhà văn tranh luận khó có thể không làm; nhưng ông đã lạm dụng; \& ông đã làm công việc đó một cách hoàn hảo. {\it Letters} của ông không được gọi là thần học. Thần học hàn lâm không phải là một khoa mà {\sc Pascal} thông thạo; khi cần thiết, các cha của Port-Royal đã đến giúp đỡ ông. {\it Letters} là tác phẩm của 1 trong những bộ óc toán học xuất sắc nhất mọi thời đại, \& của một người đàn ông của thế giới đã giải quyết, không phải các nhà thần học, mà là thế giới nói chung -- tất cả những người có học \& nhiều người ít học của giáo dân Pháp; \& với công chúng này, họ đã đạt được thành công đáng kinh ngạc.
		
		During this time {\sc Pascal} never wholly abandoned his scientific interests. Though in his religious writings he composed slowly \& painfully, \& revised often, in matters of mathematics his mind seemed to move with consummate natural ease \& grace. Discoveries \& inventions sprang from his brain without effort; among the minor devices of this later period, the 1st omnibus service in Paris is said to owe its origin to this inventiveness. But rapidly failing health, \& absorption in the great work he had in mind, left him little time \& energy during the last 2 years of his life.
		
		-- Trong thời gian này, {\sc Pascal} không bao giờ từ bỏ hoàn toàn sở thích khoa học của mình. Mặc dù trong các tác phẩm tôn giáo của mình, ông đã biên soạn chậm \& đau đớn, \& thường xuyên sửa đổi, nhưng trong các vấn đề toán học, tâm trí ông dường như chuyển động với sự thoải mái tự nhiên \& duyên dáng tuyệt đối. Những khám phá \& phát minh nảy sinh từ bộ não của ông mà không cần nỗ lực; trong số những thiết bị nhỏ của giai đoạn sau này, dịch vụ xe buýt đầu tiên ở Paris được cho là có nguồn gốc từ sự sáng tạo này. Nhưng sức khỏe suy yếu nhanh chóng, \& sự tập trung vào công trình lớn mà ông ấp ủ, khiến ông không còn nhiều thời gian \& năng lượng trong 2 năm cuối đời.
		
		The plan of what we call the {\it Pens\'ees} formed itself about 1660. The completed book was to have been a carefully constructed defence of Christianity, a true Apology \& a kind of Grammar of Assent, setting forth the reasons which will convince the intellect. As I have indicated before, {\sc Pascal} was not a theologian, \& on dogmatic theology had resource to his spiritual advisers. Nor was he indeed a systematic philosopher. He was a man with an immense genius for science, \& at the same time a natural psychologist \& moralist. As he was a great literary artist, his book would have been also his own spiritual autobiography; his style, free from all diminishing idiosyncrasies, was yet very personal. Above all, he was a man of strong passions; \& his intellectual passion for truth was reinforced by his passionate dissatisfaction with human life unless a spiritual explanation could be found.
		
		-- Kế hoạch của cái mà chúng ta gọi là {\it Pens\'ees} hình thành vào khoảng năm 1660. Cuốn sách hoàn thành là một sự bảo vệ được xây dựng cẩn thận cho Cơ đốc giáo, một Lời biện hộ thực sự \& một loại Ngữ pháp của Sự đồng thuận, nêu ra những lý do sẽ thuyết phục được trí tuệ. Như tôi đã chỉ ra trước đây, {\sc Pascal} không phải là một nhà thần học, \& về thần học giáo điều đã có nguồn lực cho các cố vấn tâm linh của mình. Ông cũng không thực sự là một triết gia có hệ thống. Ông là một người có thiên tài khoa học to lớn, \& đồng thời là một nhà tâm lý học tự nhiên \& nhà đạo đức. Vì ông là một nghệ sĩ văn học vĩ đại, cuốn sách của ông cũng sẽ là cuốn tự truyện tâm linh của riêng ông; phong cách của ông, không có bất kỳ đặc điểm riêng nào đang suy giảm, nhưng lại rất riêng tư. Trên hết, ông là một người có niềm đam mê mạnh mẽ; \& niềm đam mê trí tuệ của ông đối với chân lý được củng cố bởi sự bất mãn mãnh liệt của ông với cuộc sống con người trừ khi có thể tìm thấy một lời giải thích về mặt tâm linh.
		
		We must regard the {\it Pens\'ees} as merely the 1st notes for a work which he left far from completion; we have, in Sainte-Beuve's words, a tower of which the stones have been laid on each other, but not cemented, \& the structure unfinished. In early years his memory had been amazingly retentive of anything that he wished to remember; \& had it not been impaired by increasing illness \& pain, he probably would not have been obliged to set down these notes at all. But taking the book as it is left to us, we still find that it occupies a unique place in the history of French literature \& in the history of religious meditation.
		
		-- Chúng ta phải coi {\it Pens\'ees} chỉ là những ghi chú đầu tiên cho một tác phẩm mà ông đã để lại rất lâu trước khi hoàn thành; chúng ta có, theo lời của Sainte-Beuve, một tòa tháp mà các viên đá đã được xếp chồng lên nhau, nhưng không được gắn xi măng, \& cấu trúc còn dang dở. Trong những năm đầu, trí nhớ của ông đã ghi nhớ một cách đáng kinh ngạc bất cứ điều gì ông muốn nhớ; \& nếu không bị suy yếu do bệnh tật ngày càng tăng \& cơn đau, có lẽ ông đã không phải ghi lại những ghi chú này. Nhưng khi xem xét cuốn sách như những gì nó được để lại cho chúng ta, chúng ta vẫn thấy rằng nó chiếm một vị trí độc đáo trong lịch sử văn học Pháp \& trong lịch sử thiền định tôn giáo.
		
		To understand the method which {\sc Pascal} employs, the reader must be prepared to follow the process of the mind of the intelligent believer. The Christian thinker -- \& I mean the man who is trying consciously \& conscientiously to explain to himself the sequence which culminated in faith, rather than the public apologist -- proceeds by rejection \& elimination. He finds the world to be so \& so; he finds its character inexplicable by any non-religious theory; among religions he finds Christianity, \& Catholic Christianity, to account most satisfactorily for the world \& especially for the moral world within; \& thus, by what {\sc Newman} calls ``powerful \& concurrent'' reasons, he finds himself inexorably committed to the dogma of the Incarnation. To the unbeliever, this method seems disingenuous \& perverse; for the unbeliever is, as a rule, not so greatly troubled to explain the world to himself, nor so greatly distressed by its disorder; nor is he generally concerned (in modern terms) to ``preserve values.'' He does not consider that if certain emotional states, certain developments of character, \& what in the highest sense can be called ``saintliness'' are inherently \& by inspection known to be good, then the satisfactory explanation of the world must be an explanation which will admit the ``reality'' of these values. Nor does he consider such reasoning admissible; he would, so to speak, trim his values according to his cloth, because to him such values are of no great value. The unbeliever starts from the other end, \& as likely as not with the question: Is a case of human partheongenesis credible? \& this he would call going straight to the heart of the matter. Now {\sc Pascal}'s method is, on the whole, the method natural \& right for the Christian; \& the opposite method is that taken by {\sc Voltaire}. It is worth while to remember that {\sc Voltaire}, in his attempt to refute {\sc Pascal}, has given once \& for all the type of such refutation; \& that later opponents of {\sc Pascal}'s Apology for the Christian Faith have contributed little beyond psychological irrelevancies. For {\sc Voltaire} has presented, better than any one since, what is the unbelieving point of view; \& in the end we must all choose for ourselves between 1 point of view \& another.
		
		-- Để hiểu phương pháp mà {\sc Pascal} sử dụng, người đọc phải chuẩn bị theo dõi quá trình của tâm trí của người tin thông minh. Nhà tư tưởng Cơ đốc giáo -- \& ý tôi là người đang cố gắng một cách có ý thức \& tận tâm để giải thích cho chính mình trình tự dẫn đến đức tin, chứ không phải là nhà biện hộ công khai -- tiến hành bằng cách từ chối \& loại bỏ. Anh ta thấy thế giới là như vậy \& vậy; anh ta thấy bản chất của nó không thể giải thích được bằng bất kỳ lý thuyết phi tôn giáo nào; trong số các tôn giáo, anh ta thấy Cơ đốc giáo, \& Cơ đốc giáo Công giáo, giải thích thỏa đáng nhất cho thế giới \& đặc biệt là cho thế giới đạo đức bên trong; \& do đó, theo những gì {\sc Newman} gọi là những lý do ``mạnh mẽ \& đồng thời'', anh ta thấy mình không thể tránh khỏi việc cam kết với giáo điều về Sự nhập thể. Đối với người không tin, phương pháp này có vẻ không chân thành \& trái ngược; vì theo quy luật, người không tin không quá bận tâm đến việc giải thích thế giới cho chính mình, cũng không quá đau khổ vì sự hỗn loạn của nó; ông cũng không quan tâm (theo thuật ngữ hiện đại) đến việc ``bảo tồn các giá trị.'' Ông không cho rằng nếu một số trạng thái cảm xúc, một số diễn biến tính cách, \& những gì theo nghĩa cao nhất có thể được gọi là ``sự thánh thiện'' vốn có \& khi kiểm tra được biết là tốt, thì lời giải thích thỏa đáng về thế giới phải là lời giải thích thừa nhận ``thực tế'' của những giá trị này. Ông cũng không coi lý luận như vậy là có thể chấp nhận được; có thể nói, ông sẽ cắt xén các giá trị của mình theo cách của mình, bởi vì với ông, những giá trị như vậy không có giá trị lớn. Người không tin bắt đầu từ đầu kia, \& có khả năng như không với câu hỏi: Trường hợp sinh sản đơn tính ở người có đáng tin không? \& ông sẽ gọi đây là đi thẳng vào trọng tâm của vấn đề. Bây giờ, phương pháp của {\sc Pascal}, nhìn chung, là phương pháp tự nhiên \& đúng đắn đối với người theo đạo Thiên chúa; \& phương pháp ngược lại là phương pháp của {\sc Voltaire}. Điều đáng ghi nhớ là {\sc Voltaire}, trong nỗ lực bác bỏ {\sc Pascal}, đã đưa ra một lần \& cho tất cả các loại bác bỏ như vậy; \& rằng những người phản đối sau này của Apology for the Christian Faith của {\sc Pascal} đã đóng góp rất ít ngoài những sự không liên quan về mặt tâm lý. Vì {\sc Voltaire} đã trình bày, tốt hơn bất kỳ ai kể từ đó, quan điểm của người không tin là gì; \& cuối cùng, tất cả chúng ta phải tự mình lựa chọn giữa 1 quan điểm \& quan điểm khác.
		
		I have said above that {\sc Pascal}'s method is ``on the whole'' that of the typical Christian apologist; \& this reservation was directed as {\sc Pascal}'s belief in miracles, which plays a larger part in his construction than it would in that, at least, of the modern liberal Catholic. It would seem fantastic to accept Christianity because we 1st believe the Gospel miracles to be true, \& it would seem impious to accept it primarily because we believe more recent miracles to be true; we accept the miracles, or some miracles, to be true because we believe the Gospel of {\sc Jesus Christ}: we found our belief in the miracles on the Gospel, not our belief in the Gospel on the miracles. But it must be remembered that {\sc Pascal} had been deeply impressed by a contemporary miracle, known as the miracle of th Holy Thorn: a thorn reputed to have been preserved from the Crown of Our Lord was pressed upon an ulcer which quickly healed. {\sc Sainte-Beuve}, who as a medical man felt himself on solid ground, discusses fully the possible explanation of this apparent miracle. It is true that the miracle happened at Port-Royal, \& that it arrived opportunely to revive the depressed spirits of the community in its political afflictions; \& it is likely that {\sc Pascal} was the more inclined to believe a miracle which was performed upon his beloved sister. In any case, it probably led him to assign a place to miracles, in his study of faith, which is not quite that which we should give to them ourselves.
		
		-- Tôi đã nói ở trên rằng phương pháp của {\sc Pascal} ``nói chung'' là phương pháp của những nhà biện giáo Cơ đốc điển hình; \& sự bảo lưu này được chỉ ra là niềm tin của {\sc Pascal} vào phép lạ, đóng vai trò lớn hơn trong quá trình xây dựng của ông so với ít nhất là trong quá trình xây dựng của người Công giáo tự do hiện đại. Có vẻ như thật kỳ lạ khi chấp nhận Cơ đốc giáo vì trước tiên chúng ta tin rằng các phép lạ trong Phúc âm là có thật, \& có vẻ như là vô đạo đức khi chấp nhận nó chủ yếu vì chúng ta tin rằng các phép lạ gần đây hơn là có thật; chúng ta chấp nhận các phép lạ, hoặc một số phép lạ, là có thật vì chúng ta tin vào Phúc âm của {\sc Jesus Christ}: chúng ta thấy niềm tin của mình vào các phép lạ dựa trên Phúc âm, chứ không phải niềm tin của chúng ta vào Phúc âm về các phép lạ. Nhưng phải nhớ rằng {\sc Pascal} đã bị ấn tượng sâu sắc bởi một phép lạ đương thời, được gọi là phép lạ của Gai Thánh: một cái gai được cho là đã được bảo quản từ Vương miện của Chúa chúng ta đã được ấn vào một vết loét và nhanh chóng lành lại. {\sc Sainte-Beuve}, với tư cách là một bác sĩ, cảm thấy mình đang ở trên một nền tảng vững chắc, đã thảo luận đầy đủ về lời giải thích có thể có của phép lạ rõ ràng này. Đúng là phép lạ đã xảy ra tại Port-Royal, \& rằng nó đã đến đúng lúc để vực dậy tinh thần chán nản của cộng đồng trong những đau khổ chính trị của mình; \& có khả năng là {\sc Pascal} có xu hướng tin nhiều hơn vào một phép lạ đã được thực hiện trên người chị gái yêu dấu của mình. Trong mọi trường hợp, có lẽ nó đã khiến ông dành một vị trí cho phép lạ, trong nghiên cứu của ông về đức tin, mà không hẳn là điều mà chúng ta nên dành cho chúng.
		
		Now the great adversary against whom {\sc Pascal} set himself, from the time of his 1st conversations with {\sc M. de Saci} at Port-Royal, was {\sc Montaigne}. 
		
		
		\item {\sf Thoughts on Mind \& on Style.}
		
		\fbox{1} {\it The difference between the mathematical \& the intuitive mind.} -- In the one the principles are palpable, but removed from ordinary use; so that for want of habit it is difficult to turn one's mind in that direction: but if one turns it thither ever so little, one sees the principles fully, \& one must have a quite inaccurate mind who reasons wrongly from principles so plain that it is almost impossible they should escape notice.
		
		-- {\it Sự khác biệt giữa toán học \& tâm trí trực quan.} -- Trong trường hợp thứ nhất, các nguyên tắc là rõ ràng, nhưng tách biệt khỏi cách sử dụng thông thường; do đó, nếu không có thói quen, rất khó để hướng tâm trí theo hướng đó: nhưng nếu hướng tâm trí theo hướng đó 1 chút, người ta sẽ thấy các nguyên tắc 1 cách đầy đủ, \& người ta hẳn phải có 1 tâm trí hoàn toàn không chính xác, người lý luận sai lầm từ các nguyên tắc quá rõ ràng đến mức gần như không thể thoát khỏi sự chú ý.
		
		But in the intuitive mind the principles are found in common use, \& are before the eyes of everybody. One has only to look, \& no effort is necessary; it is only a question of good eyesight, but it must be good, for the principles are so subtle \& so numerous, that it is almost impossible but that some escape notice. Now the omission of 1 principle leads to error; thus one must have very clear sight to see all the principles, \& in the next place an accurate mind not to draw false deductions from known principles.
		
		-- Nhưng trong tâm trí trực giác, các nguyên tắc được tìm thấy trong cách sử dụng phổ biến, \& nằm trước mắt mọi người. Người ta chỉ cần nhìn, \& không cần nỗ lực; chỉ là vấn đề về thị lực tốt, nhưng phải tốt, vì các nguyên tắc rất tinh tế \& rất nhiều, đến nỗi gần như không thể không có 1 số người thoát khỏi sự chú ý. Bây giờ việc bỏ sót 1 nguyên tắc dẫn đến sai lầm; do đó, người ta phải có thị lực rất rõ ràng để thấy tất cả các nguyên tắc, \& tiếp theo là 1 tâm trí chính xác để không rút ra những suy luận sai lầm từ các nguyên tắc đã biết.
				
		All mathematicians would then be intuitive if they had clear sight, for they do not reason incorrectly from principles known to them; \& intuitive minds would be mathematical if they could turn their eyes to the principles of mathematics to which they are unused.
		
		-- Tất cả các nhà toán học sẽ có trực giác nếu họ có tầm nhìn rõ ràng, vì họ không lý luận sai từ các nguyên tắc mà họ biết; \& những tâm trí trực giác sẽ có tư duy toán học nếu họ có thể hướng mắt đến các nguyên tắc toán học mà họ chưa quen.
		
		The reason, therefore, that some intuitive minds are not mathematical is that they cannot at all turn their attention to the principles of mathematics. But the reason that mathematicians are not intuitive is that they do not see what is before them, \& that, accustomed to the exact \& plain principles of mathematics, \& not reasoning till they have well inspected \& arranged their principles, they are lost in matters of intuition where the principles do not allow of such arrangement. They are scarcely seen; they are felt rather than seen; there is the greatest difficulty in making them felt by those who do not of themselves perceive them. These principles are so fine \& so numerous that a very delicate \& very clear sense is needed to perceive them, \& to judge rightly \& justly when they are perceived, without for the most part being able to demonstrate them in order as in mathematics; because the principles are not known to us in the same way, \& because it would be an endless matter to undertake it. We must see the matter at once, at 1 glance, \& not by a process of reasoning, at least to a certain degree. \& thus it is rare that mathematicians are intuitive, \& that men of intuition are mathematicians, because mathematicians wish to treat matters of intuition mathematically, \& make themselves ridiculous, wishing to begin with definitions \& then with axioms, which is not the way to proceed in this kind of reasoning. Not that the mind does not do so, but it does it tacitly, naturally, \& without technical rules; for the expression of it is beyond all men, \& only a few can feel it.
		
		-- Do đó, lý do tại sao 1 số tâm trí trực quan không phải là toán học là vì họ không thể hướng sự chú ý của mình vào các nguyên tắc toán học. Nhưng lý do tại sao các nhà toán học không trực quan là vì họ không nhìn thấy những gì trước mắt họ, \& rằng, quen với các nguyên tắc toán học chính xác \& rõ ràng, \& không lý luận cho đến khi họ kiểm tra kỹ \& sắp xếp các nguyên tắc của mình, họ bị lạc vào các vấn đề trực quan mà các nguyên tắc không cho phép sắp xếp như vậy. Chúng hầu như không được nhìn thấy; chúng được cảm nhận hơn là được nhìn thấy; có khó khăn lớn nhất trong việc khiến chúng cảm nhận được bởi những người không tự mình nhận thức được chúng. Những nguyên tắc này rất tinh tế \& rất nhiều đến nỗi cần có 1 giác quan rất tinh tế \& rất rõ ràng để nhận thức chúng, \& để phán đoán đúng \& công bằng khi chúng được nhận thức, mà phần lớn không thể chứng minh chúng theo thứ tự như trong toán học; bởi vì chúng ta không biết các nguyên tắc theo cùng 1 cách, \& bởi vì sẽ là 1 vấn đề vô tận để thực hiện nó. Chúng ta phải nhìn thấy vấn đề ngay lập tức, chỉ trong nháy mắt, \& không phải bằng 1 quá trình lý luận, ít nhất là ở 1 mức độ nhất định. \& do đó hiếm khi các nhà toán học có trực giác, \& những người có trực giác là nhà toán học, bởi vì các nhà toán học muốn xử lý các vấn đề trực giác theo phương pháp toán học, \& làm cho mình trở nên lố bịch, muốn bắt đầu bằng các định nghĩa \& sau đó bằng các tiên đề, đó không phải là cách tiến hành trong loại lý luận này. Không phải là tâm trí không làm như vậy, nhưng nó làm điều đó 1 cách ngầm hiểu, tự nhiên, \& không có các quy tắc kỹ thuật; vì cách diễn đạt của nó vượt quá tất cả mọi người, \& chỉ 1 số ít có thể cảm nhận được.
		
		Intuitive minds, on the contrary, being thus accustomed to judge at a single glance, are so astonished when they are presented with propositions of which they understand nothing, \& the way to which is through definitions \& axioms so sterile, \& which they are not accustomed to see thus in detail, that they are repelled \& disheartened.
		
		-- Ngược lại, những tâm trí trực giác, vốn quen với việc phán đoán chỉ qua 1 cái nhìn, sẽ rất ngạc nhiên khi họ được trình bày những mệnh đề mà họ không hiểu gì cả, \& cách tiếp cận thông qua các định nghĩa \& tiên đề vô bổ, \& mà họ không quen xem xét chi tiết như vậy, đến nỗi họ cảm thấy khó chịu \& nản lòng.
		
		But dull minds are never either intuitive or mathematical.
		
		-- Nhưng những tâm trí chậm chạp không bao giờ có trực giác hoặc khả năng toán học.
		
		Mathematicians who are only mathematicians have exact minds, provided all things are explained to them by means of definitions \& axioms; otherwise they are inaccurate \& insufferable, for they are only right when the principles are quite clear.
		
		-- Các nhà toán học thực chất là các nhà toán học có tư duy chính xác, miễn là mọi thứ được giải thích cho họ thông qua các định nghĩa \& tiên đề; nếu không, chúng sẽ không chính xác \& khó chịu, vì chúng chỉ đúng khi các nguyên tắc thực sự rõ ràng.
		
		\& men of intuition who are only intuitive cannot have the patience to reach to 1st principles of things speculative \& conceptual, which they have never seen in the world, \& which are altogether out of the common.
		
		-- \& những người trực giác chỉ có trực giác không thể đủ kiên nhẫn để đạt tới những nguyên lý đầu tiên của những thứ mang tính suy đoán \& khái niệm, những thứ mà họ chưa từng thấy trên thế giới, \& hoàn toàn khác thường.
		
		\fbox{2} There are different kinds of right understanding; some have right understanding in a certain order of things, \& not in others, where they go astray. Some draw conclusions well from a few premises, \& this displays an acute judgment.
		
		-- Có nhiều loại hiểu biết đúng đắn khác nhau; 1 số hiểu biết đúng đắn theo 1 trật tự nhất định của sự vật, \& không hiểu đúng đắn theo những trật tự khác, nơi họ đi lạc. Một số rút ra kết luận tốt từ 1 vài tiền đề, \& điều này thể hiện sự phán đoán sắc sảo.
		
		Others draw conclusions well where there are many premises.
		
		-- Những người khác đưa ra kết luận tốt khi có nhiều tiền đề.
		
		E.g., the former easily learn hydrostatics, where the premises are few, but the conclusions are so fine that only the greatest acuteness can reach them.
		
		-- Ví dụ, người đầu tiên dễ dàng học được thủy tĩnh học, trong đó có ít tiền đề nhưng kết luận lại rất tinh tế đến mức chỉ có sự nhạy bén nhất mới có thể đạt tới.
		
		\& in spite of that these persons would perhaps not be great mathematicians, because mathematics contain a great number of premises, \& there is perhaps a kind of intellect that can search with ease a few premises to the bottom, \& cannot in the least penetrate those matters in which they are many premises.
		
		-- \& mặc dù những người này có lẽ không phải là những nhà toán học vĩ đại, bởi vì toán học chứa đựng rất nhiều tiền đề, \& có lẽ có 1 loại trí tuệ có thể dễ dàng tìm kiếm 1 vài tiền đề đến tận cùng, \& không thể thâm nhập vào những vấn đề có nhiều tiền đề.
		
		There are then 2 kinds of intellect: the one able to penetrate acutely \& deeply into the conclusions of given premises, \& this is the precise intellect; the other able to comprehend a great number of premises without confusing them, \& this is the mathematical intellect. The one has force \& exactness, the other comprehension. Now the one quality can exist without the other; the intellect can be strong \& narrow, \& can also be comprehensive \& weak.
		
		-- Sau đó có 2 loại trí tuệ: loại có khả năng thâm nhập sâu sắc \& vào kết luận của các tiền đề đã cho, \& đây là trí tuệ chính xác; loại còn lại có khả năng hiểu được 1 số lượng lớn các tiền đề mà không làm chúng nhầm lẫn, \& đây là trí tuệ toán học. Loại này có sức mạnh \& chính xác, loại kia có khả năng hiểu biết. Bây giờ, 1 phẩm chất có thể tồn tại mà không có phẩm chất kia; trí tuệ có thể mạnh \& hẹp, \& cũng có thể toàn diện \& yếu.
		
		\fbox{3} Those who are accustomed to judge by feeling do not understand the process of reasoning, for they would understand at 1st sight, \& are not used to seek for principles. \& others, on the contrary, who are accustomed to reason from principles, do not at all understand matters of feeling, seeking principles, \& being unable to see at a glance.
		
		-- Những người quen phán đoán bằng cảm tính không hiểu được quá trình lý luận, vì họ có thể hiểu ngay từ cái nhìn đầu tiên, \& không quen tìm kiếm nguyên tắc. \& ngược lại, những người khác quen lý luận theo nguyên tắc, hoàn toàn không hiểu vấn đề về cảm tính, tìm kiếm nguyên tắc, \& không thể nhìn thấy ngay trong cái nhìn thoáng qua.
		
		\fbox{4} {\it Mathematics, intuition.} -- True eloquence makes light of eloquence, true morality makes light of morality; i.e., the morality of the judgment, which has no rules, makes light of the morality of the intellect.
		
		-- {\it Toán học, trực giác.} -- Sự hùng biện thực sự coi thường sự hùng biện, đạo đức thực sự coi thường đạo đức; tức là đạo đức của sự phán đoán, không có quy tắc, coi thường đạo đức của trí tuệ.
		
		For it is to judgment that perception belongs, as science belongs to intellect. Intuition is the part of judgment, mathematics of intellect.
		
		-- Vì nhận thức thuộc về phán đoán, cũng như khoa học thuộc về trí tuệ. Trực giác là 1 phần của phán đoán, toán học của trí tuệ.
		
		To make light of philosophy is to be a true philosopher.
		
		-- Coi thường triết học mới là 1 triết gia thực thụ.
		
		\fbox{5} Those who judge of a work by rule are in regard to others as those who have a watch are in regard to others. One says, ``It is 2 hours ago''; the other says, ``It is only 3-quarters of an hour.'' I look at my watch, \& say to the one, ``You are weary,'' \& to the other, ``Time gallops with you''; for it is only $1.5$ hour ago, \& I laugh at those who tell me that time goes slowly with me, \& that I judge by imagination. They do not know that I judge by my watch.
		
		-- Những người đánh giá 1 tác phẩm theo quy tắc thì liên quan đến những người khác như những người có đồng hồ liên quan đến những người khác. Một người nói, ``Đã 2 giờ rồi''; người kia nói, ``Mới chỉ 3 phần tư giờ.'' Tôi nhìn vào đồng hồ của mình, \& nói với người này, ``Bạn mệt rồi,'' \& với người kia, ``Thời gian phi nước đại với bạn''; vì mới chỉ 1,5 giờ trước, \& Tôi cười những người nói với tôi rằng thời gian trôi chậm với tôi, \& rằng tôi đánh giá bằng trí tưởng tượng. Họ không biết rằng tôi đánh giá bằng đồng hồ của mình.
		
		\fbox{6} Just as we harm the understanding, we harm the feelings also.
		
		-- Khi chúng ta làm tổn hại đến sự hiểu biết, chúng ta cũng làm tổn hại đến cảm xúc.
		
		The understanding \& the feelings are moulded by intercourse; the understanding \& feelings are corrupted by intercourse. Thus good or bad society improves or corrupts them. It is, then, all-important to know how to choose in order to improve \& not to corrupt them; \& we cannot make this choice, if they be not already improved \& not corrupted. Thus a circle is formed, \& those are fortunate who escape it.
		
		-- Sự hiểu biết \& cảm xúc được định hình bởi giao hợp; sự hiểu biết \& cảm xúc bị làm hư hỏng bởi giao hợp. Do đó, xã hội tốt hay xấu cải thiện hoặc làm hư hỏng chúng. Do đó, điều quan trọng nhất là biết cách lựa chọn để cải thiện \& không làm hư hỏng chúng; \& chúng ta không thể đưa ra lựa chọn này, nếu chúng chưa được cải thiện \& không bị hư hỏng. Do đó, 1 vòng tròn được hình thành, \& những người may mắn thoát khỏi nó.
		
		\fbox{7} The greater intellect one has, the more originality one finds in men. Ordinary persons find no difference between men.
		
		-- Người ta càng có trí tuệ cao thì càng thấy tính độc đáo ở con người. Người bình thường không thấy có sự khác biệt giữa con người.
		
		\fbox{8} There are many people who listen to a sermon in the same way as they listen to vespers.
		
		-- Có nhiều người lắng nghe bài giảng theo cách tương tự như lắng nghe kinh chiều.
		
		\fbox{9} When we wish to correct with advantage, \& to show another that he errs, we must notice from what side he views the matter, for on that side it is usually true, \& admit that truth to him, but reveal to him the side on which it is false. He is satisfied with that, for he sees that he was not mistaken, \& that he only failed to see all sides. Now, no one is offended at not seeing everything; but one does not like to be mistaken, \& that perhaps arises from the fact that man naturally cannot see everything, \& that naturally he cannot err in the side he looks at, since the perceptions of our senses are always true.
		
		-- Khi chúng ta muốn sửa chữa 1 cách có lợi, \& để chỉ cho người khác thấy rằng anh ta sai, chúng ta phải để ý xem anh ta nhìn nhận vấn đề từ phía nào, vì ở phía đó thường là đúng, \& thừa nhận sự thật đó với anh ta, nhưng tiết lộ cho anh ta biết phía mà nó sai. Anh ta hài lòng với điều đó, vì anh ta thấy rằng anh ta không nhầm, \& rằng anh ta chỉ không nhìn thấy mọi phía. Bây giờ, không ai bị xúc phạm vì không nhìn thấy mọi thứ; nhưng người ta không thích bị nhầm, \& điều đó có lẽ phát sinh từ thực tế là con người tự nhiên không thể nhìn thấy mọi thứ, \& rằng tự nhiên anh ta không thể sai lầm ở phía mà anh ta nhìn vào, vì nhận thức của các giác quan của chúng ta luôn đúng.
		
		\fbox{10} People are generally better persuaded by the reasons which they have themselves discovered than by those which have come into the mind of others.
		
		-- Mọi người thường dễ bị thuyết phục hơn bởi những lý do mà họ tự khám phá ra hơn là những lý do xuất hiện trong tâm trí của người khác.
		
		\fbox{11} All great amusements are dangerous to the Christian life; but among all those which the world has invented there is none more to be feared than the theater. It is a representation of the passions so natural \& so delicate that it excites them \& gives birth to them in our hearts, \&, above all, to that of love, principally when it is represented as very chaste \& virtuous. For the more innocent it appears to innocent souls, the more they are likely to be touched by it. Its violence pleases our self-love, which immediately forms a desire to produce the same effects which are seen so well represented; \&, at the same time, we make ourselves a conscience founded on the propriety of the feelings which we see there, by which the fear of pure souls is removed, since they imagin that it cannot hurt their purity to love with a love which seems to them so reasonable.
		
		-- Mọi trò tiêu khiển lớn đều nguy hiểm cho đời sống Cơ đốc; nhưng trong số tất cả những trò mà thế gian đã phát minh ra, không có trò nào đáng sợ hơn trò kịch. Nó là sự thể hiện của những đam mê rất tự nhiên \& rất tinh tế đến nỗi nó kích thích chúng \& sinh ra chúng trong trái tim chúng ta, \& trên hết, là tình yêu, chủ yếu là khi nó được thể hiện là rất trong sáng \& đức hạnh. Vì nó càng có vẻ ngây thơ đối với những tâm hồn ngây thơ, thì chúng càng có khả năng bị nó chạm đến. Sự bạo lực của nó làm thỏa mãn lòng tự ái của chúng ta, điều này ngay lập tức hình thành mong muốn tạo ra những hiệu ứng giống như được thể hiện rất rõ ràng; \&, đồng thời, chúng ta tự tạo cho mình 1 lương tâm dựa trên sự phù hợp của những cảm xúc mà chúng ta thấy ở đó, nhờ đó nỗi sợ hãi của những tâm hồn trong sáng được loại bỏ, vì họ tưởng rằng không thể làm tổn thương sự trong sáng của họ khi yêu bằng 1 tình yêu mà đối với họ có vẻ hợp lý như vậy.
		
		So we depart from the theater with our heart so filled with all the beauty \& tenderness of love, the soul \& the mind so persuaded of its innocence, that we are quite ready to receive its 1st impressions, or rather to seek an opportunity of awakening them in the heart of another, in order that we may receive the same pleasures \& the same sacrifices which we have seen so well represented in the theater.
		
		-- Vì vậy, chúng ta rời khỏi nhà hát với trái tim tràn ngập vẻ đẹp \& sự dịu dàng của tình yêu, tâm hồn \& trí óc tin chắc vào sự ngây thơ của mình, đến nỗi chúng ta hoàn toàn sẵn sàng đón nhận những ấn tượng đầu tiên, hay đúng hơn là tìm kiếm cơ hội đánh thức chúng trong trái tim của người khác, để chúng ta có thể nhận được những thú vui \& những hy sinh giống như những gì chúng ta đã thấy được thể hiện rất tốt trong nhà hát.
		
		\fbox{12} Scaramouch, who only thinks of 1 thing.
		
		-- Scaramouch chỉ nghĩ đến 1 điều.
		
		The doctor, who speaks for 15 mins after he has said everything, so full is he of the desire of talking.
		
		-- Vị bác sĩ, người nói chuyện thêm 15 phút sau khi đã nói hết mọi thứ, vì ông ấy rất muốn được nói.
		
		\fbox{13} One likes to see the error, the passion of Cleobuline, because she is unconscious of it. She would be displeasing, if she were not deceived.
		
		-- Người ta thích nhìn thấy lỗi lầm, sự đam mê của Cleobuline, vì cô ấy không ý thức được điều đó. Cô ấy sẽ rất khó chịu nếu không bị lừa dối.
		
		\fbox{14} When a natural discourse paints a passion or an effect, one feels within oneself the truth of what one reads, which were there before, although one did not know it. Hence one is inclined to love him who makes us feel it, for he has not shown us his own riches, but ours. \& thus this benefit renders him pleasing to us, besides that such community of intellect as we have with him necessarily inclines the heart to love.
		
		-- Khi 1 bài diễn thuyết tự nhiên vẽ nên 1 niềm đam mê hay 1 hiệu ứng, người ta cảm thấy trong chính mình sự thật của những gì mình đọc, vốn đã có trước đó, mặc dù người ta không biết điều đó. Do đó, người ta có xu hướng yêu người khiến chúng ta cảm thấy điều đó, vì người đó không cho chúng ta thấy sự giàu có của riêng mình, mà là của chúng ta. \& do đó, lợi ích này khiến người đó làm chúng ta hài lòng, bên cạnh đó, cộng đồng trí tuệ như chúng ta có với người đó nhất thiết khiến trái tim hướng đến tình yêu.
		
		\fbox{15} Eloquence, which persuades by sweetness, not by authority; as a tyrant, not as a king.
		
		-- Sự hùng biện, thuyết phục bằng sự ngọt ngào, không phải bằng uy quyền; như 1 bạo chúa, không phải như 1 vị vua.
		
		\fbox{16} Eloquence is an art of saying things in such a way -- (1) that those to whom we speak may listen to them without pain \& with pleasure; (2) that they feel themselves interested, so that self-love leads them more willingly to reflection upon it.
		
		-- Sự hùng biện là nghệ thuật diễn đạt mọi việc theo cách -- (1) để những người mà chúng ta nói chuyện có thể lắng nghe mà không đau đớn \& với sự thích thú; (2) để họ cảm thấy mình quan tâm, để tình yêu bản thân khiến họ sẵn sàng suy ngẫm về điều đó hơn.
		
		It consists, then, in a correspondence which we seek to establish between the head \& the heart of those to whom we speak on the 1 hand, \& on the other, between the thoughts \& the expressions which we employ. This assumes that we have studied well the heart of man so as to know all its powers, \& then to find the just proportions of the discourse which we wish to adapt to them. We must put ourselves in the place of those who are to hear us, \& make trial on our own heart of the turn which we give to our discourse in order to see whether one is made for the other, \& whether we can assure ourselves that the hearer will be, as it were, forced to surrender. We ought to restrict ourselves, so far as possible, to the simple \& natural, \& not to magnify that which is little, or belittle that which is great. It is not enough that a thing be beautiful; it must be suitable to the subject, \& there must be in it nothing of excess or defect.
		
		-- Sau đó, nó bao gồm 1 sự tương ứng mà chúng ta tìm cách thiết lập giữa đầu \& trái tim của những người mà chúng ta nói chuyện với 1 mặt, \& mặt khác, giữa những suy nghĩ \& những biểu hiện mà chúng ta sử dụng. Điều này giả định rằng chúng ta đã nghiên cứu kỹ lưỡng trái tim của con người để biết tất cả các sức mạnh của nó, \& sau đó tìm ra tỷ lệ chính xác của bài phát biểu mà chúng ta muốn điều chỉnh cho họ. Chúng ta phải đặt mình vào vị trí của những người sẽ nghe chúng ta, \& thử nghiệm trên trái tim của chính mình về lượt mà chúng ta dành cho bài phát biểu của mình để xem liệu người này có dành cho người kia không, \& liệu chúng ta có thể tự đảm bảo rằng người nghe sẽ, như thể, bị buộc phải đầu hàng. Chúng ta nên hạn chế bản thân, trong khả năng có thể, vào những điều đơn giản \& tự nhiên, \& không phóng đại những gì nhỏ bé, hoặc hạ thấp những gì vĩ đại. Một điều gì đó đẹp thôi là chưa đủ; nó phải phù hợp với chủ đề, \& không được có bất cứ điều gì dư thừa hay khiếm khuyết trong đó.
		
		\fbox{17} Rivers are roads which move, \& which carry us whether we desire to go.
		
		-- Sông là những con đường chuyển động, \& đưa chúng ta đi dù chúng ta có muốn đi hay không.
		
		\fbox{18} When we do not know the truth of a thing, it is of advantage that there should exist a common error which determines the mind of man, as, e.g., the moon, to which is attributed the change of reasons, the progress of diseases, etc. For the chief malady of man is restless curiosity about things which he cannot understand; \& it is not so bad for him to be in error as to be curious to no purpose.
		
		-- Khi chúng ta không biết sự thật của 1 điều gì đó, thì sẽ có lợi nếu tồn tại 1 lỗi chung quyết định tâm trí con người, chẳng hạn như mặt trăng, được cho là nguyên nhân gây ra sự thay đổi lý trí, sự tiến triển của bệnh tật, v.v. Vì căn bệnh chính của con người là sự tò mò không ngừng về những thứ mà họ không thể hiểu; \& không tệ khi họ mắc lỗi bằng việc tò mò không có mục đích.
		
		The manner in which {\sc Epictetus, Montaigne, \& Salomon de Tultie} wrote, is the most usual, the most suggestive, the most remembered, \& the oftenest quoted; because it is entirely composed of thoughts born from the common talk of life. As when we speak of the common error which exists among men that the moon is the cause of everything, we never fail to say that {\sc Salomon de Tultie} says that when we do not know the truth of a thing, it is of advantage that there should exists a common error, etc.; which is the thought above.
		
		-- Cách mà {\sc Epictetus, Montaigne, \& Salomon de Tultie} viết, là cách thông thường nhất, gợi ý nhất, được nhớ đến nhiều nhất, \& được trích dẫn nhiều nhất; bởi vì nó hoàn toàn bao gồm những suy nghĩ nảy sinh từ cuộc nói chuyện chung của cuộc sống. Giống như khi chúng ta nói về lỗi chung tồn tại giữa con người rằng mặt trăng là nguyên nhân của mọi thứ, chúng ta không bao giờ không nói rằng {\sc Salomon de Tultie} nói rằng khi chúng ta không biết sự thật của 1 điều gì đó, thì việc tồn tại 1 lỗi chung là có lợi, v.v.; đó là suy nghĩ ở trên.
		
		\fbox{19} The last thing one settles in writing a book is what one should put in 1st.
		
		-- Điều cuối cùng mà người ta cần lưu ý khi viết 1 cuốn sách là nên đưa những gì vào đầu tiên.
		
		\fbox{20} {\it Order.} -- Why should I undertake to divide my virtues into 4 rather than into 6? Why should I rather establish virtue in 4, in 2, in 1? Why into {\it Abstine et sustine} rather than into ``Follow Nature,'' or ``Conduct your private affairs without injustice,'' as {\sc Plato}, or anything else? But there, you will say, everything is contained in 1 word. yes, but it is useless without explanation, \& when we come to explain it, as soon as we unfold this maxim which contains all the rest, they emerge in that 1st confusion which you desired to avoid. So, when they are all included in one, they are hidden \& useless, as in a chest, \& never appear save in their natural confusion. Nature has established them all without including one in the other.
		
		-- {\it Thứ tự.} -- Tại sao tôi phải chia đức hạnh của mình thành 4 thay vì thành 6? Tại sao tôi lại phải thiết lập đức hạnh thành 4, thành 2, thành 1? Tại sao thành {\it Abstine et sustine} thay vì thành ``Tuân theo Tự nhiên'', hay ``Tiến hành công việc riêng tư của bạn mà không có bất công'', như {\sc Plato}, hay bất cứ điều gì khác? Nhưng ở đó, bạn sẽ nói, mọi thứ đều nằm trong 1 từ. đúng, nhưng sẽ vô ích nếu không có lời giải thích, \& khi chúng ta bắt đầu giải thích, ngay khi chúng ta mở ra câu châm ngôn này chứa đựng tất cả phần còn lại, chúng xuất hiện trong sự nhầm lẫn đầu tiên mà bạn muốn tránh. Vì vậy, khi tất cả chúng được bao gồm trong một, chúng bị ẩn \& vô dụng, như trong 1 cái rương, \& không bao giờ xuất hiện ngoại trừ trong sự nhầm lẫn tự nhiên của chúng. Tự nhiên đã thiết lập tất cả chúng mà không bao gồm chúng trong nhau.
		
		\fbox{21} Nature has made all her truths independent of 1 another. Our art makes one dependent on the other. But this is not natural. Each keeps its own place.
		
		-- Thiên nhiên đã tạo ra tất cả chân lý của mình độc lập với nhau. Nghệ thuật của chúng ta làm cho cái này phụ thuộc vào cái kia. Nhưng điều này không tự nhiên. Mỗi cái giữ vị trí riêng của nó.
		
		\fbox{22} Let no one say that I have said nothing new; the arrangement of the subject is new. When we play tennis, we both play with the same ball, but 1 of us places it better.
		
		-- Đừng ai nói rằng tôi không nói điều gì mới; cách sắp xếp chủ đề là mới. Khi chúng ta chơi quần vợt, cả hai chúng ta đều chơi với cùng 1 quả bóng, nhưng 1 trong chúng ta đặt nó tốt hơn.
		
		I had as soon it said that I used words employed before. \& in the same way if the same thoughts in a different arrangement do not form a different discourse, no more do the same words in their different arrangement form different thoughts!
		
		-- Tôi đã từng nói rằng tôi sử dụng những từ đã dùng trước đó. \& tương tự như vậy, nếu cùng 1 ý nghĩ được sắp xếp khác nhau không tạo thành 1 diễn ngôn khác nhau, thì cùng 1 từ được sắp xếp khác nhau cũng không tạo thành những ý nghĩ khác nhau!		
		
		\fbox{23} Words differently arranged have a different meaning, \& meanings differently arranged have different effects.
		
		-- Các từ được sắp xếp khác nhau sẽ mang lại ý nghĩa khác nhau, \& các ý nghĩa được sắp xếp khác nhau sẽ có hiệu quả khác nhau.
		
		\fbox{24} {\it Language.} -- We should not turn the mind from 1 thing to another, except for relaxation, \& that when it is necessary \& the time suitable, \& not otherwise. For he that relaxes out of season wearies, \& he who wearies us out of season makes us languid, since we turn quite away. So much does our perverse lust like to do the contrary of what those wish to obtain from us without giving us pleasure, the coin for which we will do whatever is wanted.
		
		-- {\it Ngôn ngữ.} -- Chúng ta không nên chuyển tâm trí từ việc này sang việc khác, ngoại trừ việc thư giãn, \& khi cần thiết \& thời điểm thích hợp, \& không được làm khác. Vì người thư giãn ngoài mùa sẽ mệt mỏi, \& người làm chúng ta mệt mỏi ngoài mùa sẽ khiến chúng ta uể oải, vì chúng ta quay đi hoàn toàn. Dục vọng đồi trụy của chúng ta thích làm ngược lại những gì những người muốn có được từ chúng ta mà không làm chúng ta vui, đồng tiền mà chúng ta sẽ làm bất cứ điều gì được yêu cầu.
		
		\fbox{25} {\it Eloquence.} -- It requires the pleasant \& the real; but the pleasant must itself be drawn from the true.
		
		-- {\it Sự hùng biện.} -- Nó đòi hỏi sự dễ chịu \& sự thực; nhưng bản thân sự dễ chịu phải được rút ra từ sự thực.
		
		\fbox{26} Eloquence is a painting of thought; \& thus those who, after having painted it, add something more, make a picture instead of a portrait.
		
		-- Sự hùng biện là bức tranh của tư tưởng; \& do đó, những người sau khi đã vẽ xong, thêm thắt điều gì đó, tạo nên 1 bức tranh thay vì 1 bức chân dung.
		
		\fbox{27} {\it Miscellaneous. Language.} -- Those who make atitheses by forcing words are like those who make false windows for symmetry. Their rule is not to speak accurately, but to make apt figures of speech.
		
		-- {\it Lặt Vặt. Ngôn ngữ.} -- Những người tạo ra những sự kiện bất thường bằng cách ép buộc từ ngữ giống như những người tạo ra những cửa sổ giả để tạo sự cân xứng. Quy tắc của họ không phải là nói chính xác, mà là tạo ra những hình ảnh ẩn dụ thích hợp.
		
		\fbox{28} Symmetry is what we see at a glance; based on the fact that there is no reason for any difference, \& based also on the face of man; whence it happens that symmetry is only wanted in breadth, not in height or depth.
		
		-- Tính đối xứng là thứ chúng ta nhìn thấy ngay từ cái nhìn đầu tiên; dựa trên thực tế là không có lý do gì cho bất kỳ sự khác biệt nào, \& cũng dựa trên khuôn mặt của con người; do đó tính đối xứng chỉ được mong muốn ở chiều rộng, không phải chiều cao hay chiều sâu.
		
		\fbox{29} When we see a natural style, we are astonished \& delighted; for we expected to see an author, \& we find a man. Whereas those who have good taste, \& who seeing a book expect to find a man, are quite surprised to find an author. {\it Plus poetice quam humane locutus es}. Those honor Nature well, who teach that she can speak on everything, even on theology.
		
		-- Khi chúng ta thấy 1 phong cách tự nhiên, chúng ta ngạc nhiên \& vui mừng; vì chúng ta mong đợi nhìn thấy 1 tác giả, \& chúng ta tìm thấy 1 người đàn ông. Trong khi những người có gu thẩm mỹ tốt, \& những người nhìn thấy 1 cuốn sách mong đợi tìm thấy 1 người đàn ông, thì khá ngạc nhiên khi tìm thấy 1 tác giả. {\it Plus poetice quam humane locutus es}. Những người tôn vinh Thiên nhiên tốt, những người dạy rằng cô ấy có thể nói về mọi thứ, thậm chí về thần học.
				
		\fbox{30} We only consult the ear because the heart is wanting. The rule is uprightness.
		
		-- Chúng ta chỉ tham khảo tai vì trái tim thiếu sự chính trực. Quy tắc là sự ngay thẳng.
		
		Beauty of omission, of judgment.
		
		-- Vẻ đẹp của sự bỏ sót, của sự phán đoán.
		
		\fbox{31} All the false beauties which we blame into Cicero have their admirers, \& in great number.
		
		-- Tất cả những vẻ đẹp giả tạo mà chúng ta đổ lỗi cho Cicero đều có rất nhiều người ngưỡng mộ.
		
		\fbox{32} There is a certain standard of grace \& beauty which consists in a certain relation between our nature, such as it is, weak or strong, \& the thing which pleases us.
		
		-- Có 1 tiêu chuẩn nhất định về sự duyên dáng \& vẻ đẹp bao gồm mối quan hệ nhất định giữa bản chất của chúng ta, dù yếu hay mạnh, \& điều làm chúng ta hài lòng.
		
		Whatever is formed according to this standard pleases us, be it house, song, discourse, verse, prose, woman, birds, rivers, trees, rooms, dress, etc. Whatever is not made according to this standard displeases those who have good taste.
		
		-- Bất cứ thứ gì được hình thành theo tiêu chuẩn này đều làm chúng ta hài lòng, dù đó là ngôi nhà, bài hát, bài diễn văn, câu thơ, văn xuôi, phụ nữ, chim chóc, sông ngòi, cây cối, phòng ốc, trang phục, v.v. Bất cứ thứ gì không được tạo ra theo tiêu chuẩn này đều làm mất lòng những người có gu thẩm mỹ tốt.
		
		\& as there is a perfect relation between a song \& a house which are made after a good model, because they are like this good model, though each after its kind; even so there is a perfect relation between things made after a bad model. Not that the bad model is unique, for there are many; but each bad sonnet, e.g., on whatever false model it is formed, is just like a woman dressed after that model.
		
		-- \& vì có 1 mối quan hệ hoàn hảo giữa 1 bài hát \& 1 ngôi nhà được tạo ra theo 1 mô hình tốt, bởi vì chúng giống như mô hình tốt này, mặc dù mỗi mô hình theo loại của nó; cũng vậy có 1 mối quan hệ hoàn hảo giữa những thứ được tạo ra theo 1 mô hình xấu. Không phải là mô hình xấu là duy nhất, vì có rất nhiều; nhưng mỗi bài sonnet xấu, ví dụ, trên bất kỳ mô hình sai lầm nào mà nó được hình thành, thì cũng giống như 1 người phụ nữ ăn mặc theo mô hình đó.
		
		Nothing makes us understand better the ridiculousness of a false sonnet than to consider nature \& the standard, \& then to imagine a woman or a house made according to that standard.
		
		-- Không gì giúp chúng ta hiểu rõ hơn sự lố bịch của 1 bài thơ sonnet giả tạo hơn là xem xét thiên nhiên \& chuẩn mực, \& rồi tưởng tượng ra 1 người phụ nữ hoặc 1 ngôi nhà được xây dựng theo chuẩn mực đó.
		
		\fbox{33} {\it Poetical beauty.} -- As we speak of poetical beauty, so ought we to speak of mathematical beauty \& medical beauty. But we do not do so; \& the reason is that we know well what is the object of mathematics, \& that it consists in proofs, \& what is the object of medicine, \& that it consists in healing. But we do not know in what grace consists, which is the object of poetry. We do not know the natural model which we ought to imitate; \& through lack of this knowledge, we have coined fantastic terms, ``The golden age,'' ``The wonder of our times,'' ``Fatal,'' etc., \& call this jargon poetical beauty.
		
		-- {\it Vẻ đẹp thơ ca.} -- Khi chúng ta nói về vẻ đẹp thơ ca, chúng ta cũng nên nói về vẻ đẹp toán học \& vẻ đẹp y khoa. Nhưng chúng ta không làm như vậy; \& lý do là chúng ta biết rõ mục đích của toán học là gì, \& rằng nó bao gồm các bằng chứng, \& mục đích của y học là gì, \& rằng nó bao gồm việc chữa bệnh. Nhưng chúng ta không biết ân sủng bao gồm điều gì, đó là mục đích của thơ ca. Chúng ta không biết mô hình tự nhiên mà chúng ta nên bắt chước; \& do thiếu kiến thức này, chúng ta đã đặt ra những thuật ngữ kỳ lạ, ``Thời đại hoàng kim'', ``Kỳ quan của thời đại chúng ta'', ``Chết chóc'', v.v., \& gọi thuật ngữ này là vẻ đẹp thơ ca.
		
		But whoever imagines a woman after this model, which consists in saying little things in big words, will see a pretty girl adorned with mirrors \& chains, at whom he will smile; because we know better wherein consists the charm of woman than the charm of verse. But those who are ignorant would admire her in this dress, \& there are many villages in which she would be taken for the queen; hence we call sonnets made after this model ``Village Queens.''
		
		-- Nhưng bất kỳ ai tưởng tượng ra một người phụ nữ theo mô hình này, bao gồm việc nói những điều nhỏ nhặt bằng những từ ngữ lớn lao, sẽ thấy một cô gái xinh đẹp được trang điểm bằng gương \& dây xích, người mà anh ta sẽ mỉm cười; bởi vì chúng ta biết rõ hơn về sự quyến rũ của phụ nữ hơn là sự quyến rũ của thơ ca. Nhưng những người thiếu hiểu biết sẽ ngưỡng mộ cô ấy trong bộ váy này, \& có nhiều ngôi làng mà cô ấy sẽ được coi là nữ hoàng; do đó chúng ta gọi những bài thơ sonnet được tạo ra theo mô hình này là ``Nữ hoàng làng.''
		
		\fbox{34} No one passes in the world as skilled in verse unless he has put up the sign of a poet, a mathematician, etc. But educated people do not want a sign, \& draw little distinction between the trade of a poet \& that of an embroiderer.
		
		-- Không ai trên thế giới này được coi là có tài làm thơ trừ khi người đó có dấu hiệu của một nhà thơ, một nhà toán học, v.v. Nhưng những người có học thức không muốn có dấu hiệu, \& không phân biệt nhiều giữa nghề của một nhà thơ \& nghề của một người thêu thùa.
		
		People of education are not called poets or mathematicians, etc.; but they are all these, \& judges of all these. No one guesses what they are. When they come into society, they talk on matters about which the rest are talking. We do not observe in them 1 quality rather than another, save when they have to make use of it. But then we remember it, for it is characteristic of such persons that we do not say of them that they are fine speakers, when it is not a question of oratory, \& that we say of them that they are fine speakers, when it is such a question.
		
		-- Những người có học thức không được gọi là nhà thơ hay nhà toán học, v.v.; nhưng họ là tất cả những điều này, \& thẩm phán của tất cả những điều này. Không ai đoán được họ là gì. Khi họ bước vào xã hội, họ nói về những vấn đề mà những người còn lại đang nói. Chúng ta không quan sát thấy ở họ phẩm chất này hơn phẩm chất khác, trừ khi họ phải sử dụng nó. Nhưng sau đó chúng ta nhớ ra điều đó, vì đặc điểm của những người như vậy là chúng ta không nói về họ rằng họ là những người nói hay, khi đó không phải là vấn đề hùng biện, \& rằng chúng ta nói về họ rằng họ là những người nói hay, khi đó là một vấn đề như vậy.
		
		It is therefore false praise to give a man when we say of him, on his entry, that he is a very clever poet; \& it is a bad sign when a man is not asked to give his judgment on some verses.
		
		-- Do đó, thật là lời khen sai lầm khi chúng ta nói về một người khi họ mới bước vào rằng họ là một nhà thơ rất thông minh; \& thật là một dấu hiệu xấu khi một người không được yêu cầu đưa ra đánh giá của mình về một số câu thơ.
		
		\fbox{35} We should not be able to say of a man, ``He is a mathematician,'' or ``a preacher,'' or ``eloquent''; but that he is ``a gentleman.'' That universal quality alone pleases me. It is a bad sign when, on seeing a person, you remember his book. I would prefer you to see no quality till you meet it \& have occasion to use it ({\it Ne quid nimis}), for fear some one quality prevail \& designate the man. Let none think him a fine speaker, unless oratory be in question, \& then let them think it.
		
		-- Chúng ta không thể nói về một người đàn ông rằng ``Anh ta là một nhà toán học'', hoặc ``một nhà thuyết giáo'', hoặc ``hùng biện''; mà là ``một quý ông''. Chỉ riêng phẩm chất phổ quát đó cũng đủ làm tôi hài lòng. Thật là một dấu hiệu xấu khi, khi nhìn thấy một người, bạn nhớ đến cuốn sách của anh ta. Tôi muốn bạn không thấy phẩm chất nào cho đến khi bạn gặp nó \& có dịp sử dụng nó ({\it Ne quid nimis}), vì sợ rằng một phẩm chất nào đó sẽ chiếm ưu thế \& chỉ định người đàn ông đó. Đừng để ai nghĩ anh ta là một diễn giả giỏi, trừ khi có vấn đề về hùng biện, \& sau đó hãy để họ nghĩ như vậy.
		
		\fbox{36} Man is full of wants: he loves only those who can satisfy them all. ``This one is a good mathematician,'' one will say. But I have nothing to do with mathematics; he would take me for a proposition. ``That one is a good soldier.'' He would take me for a besieged town. I need, then, an upright man who can accommodate himself generally to all my wants.
		
		-- Con người đầy những nhu cầu: anh ta chỉ yêu những người có thể thỏa mãn tất cả những nhu cầu đó. ``Người này là một nhà toán học giỏi,'' người ta sẽ nói. Nhưng tôi không liên quan gì đến toán học; anh ta sẽ coi tôi là một đề xuất. ``Người kia là một người lính giỏi.'' Anh ta sẽ coi tôi là một thị trấn bị bao vây. Vậy thì, tôi cần một người đàn ông ngay thẳng có thể đáp ứng mọi nhu cầu của tôi.
		
		\fbox{37} [Since we cannot be universal \& know all that is to be known of everything, we ought to know a little about everything. For it is far better to know something about everything than to know all about 1 thing. This universality is the best. If we can have both, still better; but if we must choose, we ought to choose the former. \& the world feels this \& does so; for the world is often a good judge.]
		
		-- Vì chúng ta không thể phổ quát \& biết tất cả những gì cần biết về mọi thứ, chúng ta nên biết một chút về mọi thứ. Vì biết một chút về mọi thứ thì tốt hơn nhiều so với biết tất cả về 1 thứ. Tính phổ quát này là tốt nhất. Nếu chúng ta có thể có cả hai, thì vẫn tốt hơn; nhưng nếu chúng ta phải lựa chọn, chúng ta nên chọn cái trước. \& thế giới cảm thấy điều này \& thực sự như vậy; vì thế giới thường là một thẩm phán tốt.
		
		\fbox{38} A poet \& not an honest man.
		
		-- Một nhà thơ \& không phải là một người trung thực.
		
		\fbox{39} If lighting fell on low places, etc., poets, \& those who can only reason about things of that kind, would lack proofs.
		
		-- Nếu ánh sáng chiếu xuống những nơi thấp, v.v., các nhà thơ \& những người chỉ có thể lý luận về những điều như thế, sẽ thiếu bằng chứng.
		
		\fbox{40} If we wished to prove the examples which we take to prove other things, we should have to take those other things to be examples; for, as we always believe the difficulty is in what we wish to prove, we find the examples clearer \& a help to demonstration.
		
		-- Nếu chúng ta muốn chứng minh các ví dụ mà chúng ta lấy để chứng minh những điều khác, chúng ta phải lấy những điều khác đó làm ví dụ; vì chúng ta luôn tin rằng khó khăn nằm ở điều chúng ta muốn chứng minh, chúng ta thấy các ví dụ rõ ràng hơn \& giúp ích cho việc chứng minh.
		
		Thus when we wish to demonstrate a general theorem, we must give the rule as applied to a particular case; but if we wish to demonstrate a particular case, we must begin with the general rule. For we always find the thing obscure which we wish to prove, \& that clear which we use for the proof; for, when a thing is put forward to be proved, we 1st fill ourselves with the imagination that it is therefore obscure, \& on the contrary that what is to prove it is clear, \& so we understand it easily.
		
		-- Vì vậy, khi chúng ta muốn chứng minh một định lý tổng quát, chúng ta phải đưa ra quy tắc áp dụng cho một trường hợp cụ thể; nhưng nếu chúng ta muốn chứng minh một trường hợp cụ thể, chúng ta phải bắt đầu với quy tắc tổng quát. Vì chúng ta luôn thấy điều mà chúng ta muốn chứng minh là mơ hồ, \& điều rõ ràng mà chúng ta sử dụng để chứng minh; vì, khi một điều được đưa ra để chứng minh, trước tiên chúng ta tự lấp đầy mình bằng trí tưởng tượng rằng do đó nó mơ hồ, \& ngược lại, điều cần chứng minh là rõ ràng, \& vì vậy chúng ta dễ hiểu.
		
		\fbox{41} {\it Epigrams of Martial.} -- Man loves malice, but not against 1-eyed men nor the unfortunate, but against the fortunate \& proud. People are mistaken in thinking otherwise.
		
		-- {\it Những câu châm ngôn về võ thuật.} -- Con người thích sự độc ác, nhưng không phải với những kẻ một mắt hay những kẻ bất hạnh, mà với những kẻ may mắn \& kiêu ngạo. Mọi người đã nhầm lẫn khi nghĩ khác.
		
		For lust is the source of all our actions, \& humanity, etc. We must please those who have humane \& tender feelings. That epigram about 2 1-eyed people is worthless, for it does not console them, \& only gives a point to the author's glory. All that is only for the sake of the author is worthless. {\it Ambitiosa recident ornamenta}.
		
		-- Vì dục vọng là nguồn gốc của mọi hành động, \& nhân tính, v.v. của chúng ta. Chúng ta phải làm hài lòng những người có tình cảm nhân đạo \& dịu dàng. Câu châm ngôn về 2 người 1 mắt là vô giá trị, vì nó không an ủi họ, \& chỉ nêu bật vinh quang của tác giả. Tất cả những gì chỉ vì lợi ích của tác giả đều vô giá trị. {\it Ambitiosa recident ornamenta}.
		
		\fbox{42} To call a king ``Prince'' is pleasing, because it diminishes his rank.
		
		-- Gọi một vị vua là ``Hoàng tử'' thì dễ chịu hơn, vì nó làm giảm địa vị của ông ta.
		
		\fbox{43} Certain authors, speaking of their works, say, ``My book,'' ``My commentary,'' ``My history,'' etc. They resemble middle-class people who have a house of their own, \& always have ``My house'' on their tongue. They would do better to say, ``Our book,'' ``Our commentary,'' ``Our history,'' etc., because there is in them usually more of other people's than their own.
		
		-- Một số tác giả, khi nói về tác phẩm của mình, nói rằng, ``Cuốn sách của tôi,'' ``Bình luận của tôi,'' ``Lịch sử của tôi,'' v.v. Họ giống như những người trung lưu có một ngôi nhà riêng, \& luôn có ``Ngôi nhà của tôi'' trên lưỡi của họ. Họ sẽ làm tốt hơn nếu nói rằng, ``Cuốn sách của chúng tôi,'' ``Bình luận của chúng tôi,'' ``Lịch sử của chúng tôi,'' v.v., vì trong đó thường có nhiều thứ của người khác hơn là của chính họ.
		
		\fbox{44} Do you wish people to believe good of you? Don't speak.
		
		-- Bạn muốn mọi người tin tốt về bạn? Đừng nói gì cả.
		
		\fbox{45} Languages are ciphers, wherein letters are not changed into letters, but words into words, so that unknown language is decipherable.
		
		-- Ngôn ngữ là mật mã, trong đó các chữ cái không được chuyển thành chữ cái mà chuyển từ từ thành từ, do đó ngôn ngữ chưa biết có thể giải mã được.
		
		\fbox{46} A maker of witticisms, a bad character.
		
		-- 1 kẻ hay nói đùa, một nhân vật xấu.
		
		\fbox{47} There are some who speak well \& write badly. For the place \& the audience warm them, \& draw from their minds more than they think of without that warmth.
		
		-- Có một số người nói hay \& viết dở. Đối với nơi \& khán giả sưởi ấm họ, \& rút ra từ tâm trí họ nhiều hơn họ nghĩ mà không có sự ấm áp đó.
		
		\fbox{48} When we find words repeated in a discourse, \&, in trying to correct them, discover that they are so appropriate that we would spoil the discourse, we must leave them alone. This is the test; \& our attempt is the work of envy, which is blind, \& does not see that repetition is not in this place a fault; for there is no general rule.
		
		-- Khi chúng ta tìm thấy những từ được lặp lại trong một bài diễn thuyết, \&, khi cố gắng sửa chúng, phát hiện ra rằng chúng quá phù hợp đến mức chúng ta có thể làm hỏng bài diễn thuyết, chúng ta phải để chúng yên. Đây là bài kiểm tra; \& nỗ lực của chúng ta là công việc của sự đố kỵ, vốn mù quáng, \& không thấy rằng sự lặp lại không phải là một lỗi ở đây; vì không có quy tắc chung nào cả.
		
		\fbox{49} To mask nature \& disguise her. No more king, pope, bishop -- but {\it agust monarch}, etc.; not Paris -- {\it the capital of the kingdom}. There are places in which we ought to call Paris, Paris, \& others in which we ought to call it the capital of the kingdom.
		
		-- Để che giấu thiên nhiên \& ngụy trang cho nó. Không còn vua, giáo hoàng, giám mục nữa -- nhưng {\it agust monarch}, v.v.; không phải Paris -- {\it là thủ đô của vương quốc}. Có những nơi mà chúng ta nên gọi là Paris, Paris, \& những nơi khác mà chúng ta nên gọi là thủ đô của vương quốc.
		
		\fbox{50} The same meaning changes with the words which express it. Meanings receive their dignity from words instead of giving it to them. Examples should be sought $\ldots$
		
		-- Ý nghĩa tương tự thay đổi theo các từ diễn đạt nó. Ý nghĩa nhận được phẩm giá của chúng từ các từ thay vì trao nó cho chúng. Cần tìm ví dụ $\ldots$
		
		\fbox{51} Sceptic, for obstinate.
		
		-- Người hoài nghi, cố chấp.
		
		\fbox{52} No one calls another a Cartesian but he who is one himself, a pedant but a pedant, a provincial but a provincial; \& I would wager it was the printer who put it on the title of {\it Letters to a Provincial}.
		
		-- Không ai gọi người khác là Descartes ngoại trừ chính người đó, một kẻ hay soi mói nhưng vẫn là kẻ hay soi mói, một người tỉnh lẻ nhưng vẫn là người tỉnh lẻ; \& Tôi cá là chính người thợ in đã đặt điều đó vào tiêu đề của {\it Thư gửi một Tỉnh Trưởng}.
		
		\fbox{53} A carriage {\it upset} or {\it overturned}, according to the meaning {\it To spread abroad} or {\it upset}, according to the meaning. (The argument by force of M. le Ma\^itre over the friar.)
		
		-- Một cỗ xe ngựa {\it bị lật} hoặc {\it bị lật đổ}, theo nghĩa là {\it lan rộng ra} hoặc {\it bị lật}, theo nghĩa. (Cuộc tranh luận bằng vũ lực của M. le Ma\^itre về tu sĩ.)
		
		\fbox{54} {\it Miscellaneous.} -- A form of speech, ``I should have liked to apply myself to that.''
		
		-- {\it Khác.} -- Một dạng lời nói, ``Tôi muốn tham gia vào việc đó.''
		
		\fbox{55} The {\it aperitive} virtue of a key, the {\it attractive} virtue of a hook.
		
		-- Tính chất khai vị của chìa khóa, tính chất hấp dẫn của lưỡi câu.
		
		\fbox{56} To guess: ``The part that I take in your trouble.'' The Cardinal did not want to be guessed.
		
		-- Để đoán: ``Vai trò của tôi trong rắc rối của anh.'' Đức Hồng y không muốn bị đoán.
		
		``My mind is disquieted.'' {\it I am disquieted} is better.
		
		-- ``Tôi đang bồn chồn.'' Nói là tôi bồn chồn thì tốt hơn. {\tt[dịch chưa sát nghĩa]}
		
		\fbox{57} I always feel uncomfortable under such compliments as these: ``I have given you a great deal of trouble,'' ``I am afraid I am boring you,'' ``I fear this is too long.'' We either carry our audience with us, or irritate them.
		
		-- Tôi luôn cảm thấy không thoải mái khi nhận được những lời khen như thế này: ``Tôi đã gây ra cho bạn rất nhiều rắc rối,'' ``Tôi sợ rằng tôi đang làm bạn chán,'' ``Tôi sợ rằng bài này quá dài.'' Chúng ta hoặc là sẽ thu hút được khán giả, hoặc sẽ làm họ khó chịu.
		
		\fbox{58} You are ungraceful: ``Excuse me, pray.'' Without that excuse I would not have known there was anything amiss. ``With reverence be it spoken $\ldots$'' The only thing bad is their excuse.
		
		-- Bạn thật vô duyên: ``Xin lỗi, cầu nguyện.'' Nếu không có lời bào chữa đó, tôi sẽ không biết có điều gì đó không ổn. ``Hãy nói một cách tôn kính $\ldots$'' Điều duy nhất tệ là lời bào chữa của họ.
		
		\fbox{59} ``To extinguish the torch of sedition''; too luxuriant. ``The restlessness of his genius''; 2 superfluous grand words.
		
		-- ``Để dập tắt ngọn đuốc của sự phản loạn''; quá xa hoa. ``Sự bồn chồn của thiên tài của ông''; 2 từ ngữ hoa mỹ thừa thãi.		
			 
		\item {\sf The Misery of Man without God.}
		
		\fbox{60} {\it1st part}: Misery of man without God.
		
		-- Phần 1: Sự khốn khổ của con người khi không có Chúa.
		
		{\it2nd part}: Happiness of man with God.
		
		-- Phần 2: Hạnh phúc của con người với Thiên Chúa.
		
		Or, {\it1st part}: That nature is corrupt. Proved by nature itself.
		
		-- Hoặc, phần 1: Rằng thiên nhiên là hư hỏng. Được chứng minh bởi chính thiên nhiên.
		
		{\it2nd part}: That there is a Redeemer. Proved by Scripture.
		
		-- Phần 2: Có một Đấng Cứu Chuộc. Được chứng minh bằng Kinh Thánh.
		
		\fbox{61} {\it Order.} -- I might well have taken this discourse in an order like this: to show the vanity of all conditions of men, to show the vanity of ordinary lives, \& then the vanity of philosophic lives, sceptics, stoics; but the order would not have been kept. I know a little what it is, \& how few people understand it. No human science can keep it. Saint Thomas did not keep it. Mathematics keep it, but they are useless on account of their depth.
		
		-- Trật tự. -- Tôi có thể đã đưa bài diễn thuyết này theo một trật tự như thế này: để chỉ ra sự phù phiếm của mọi điều kiện của con người, để chỉ ra sự phù phiếm của cuộc sống bình thường, \& sau đó là sự phù phiếm của cuộc sống triết học, những người hoài nghi, những người khắc kỷ; nhưng trật tự sẽ không được duy trì. Tôi biết một chút về nó, \& rất ít người hiểu được nó. Không có khoa học nào của con người có thể duy trì nó. Thánh Thomas đã không duy trì nó. Toán học duy trì nó, nhưng chúng vô dụng vì độ sâu của chúng.
		
		\fbox{62} {\it Preface to the 1st part.} -- To speak of those who have treated of the knowledge of self; of the divisions of Charron, which sadden \& weary us; of the confusion of Montaigne; that he was quite aware of his want of method, \& shunned it by jumping from subject to subject; that he sought to be fashionable.
		
		-- {\it Lời tựa cho phần 1.} -- Để nói về những người đã đề cập đến kiến thức về bản thân; về sự chia rẽ của Charron, khiến chúng ta buồn \& mệt mỏi; về sự nhầm lẫn của Montaigne; rằng ông hoàn toàn nhận thức được sự thiếu phương pháp của mình, \& tránh nó bằng cách nhảy từ chủ đề này sang chủ đề khác; rằng ông tìm cách trở nên hợp thời trang.
		
		His foolish project of describing himself! \& this not casually \& against his maxims, since every one makes mistakes, but by his maxims themselves, \& by 1st \& chief design. For to say silly things by chance \& weakness is a common misfortune; but to say them intentionally is intolerable; \& to say such as that $\ldots$
		
		-- Dự án ngu ngốc của anh ta là tự mô tả bản thân! \& điều này không phải ngẫu nhiên \& chống lại các châm ngôn của anh ta, vì mọi người đều mắc lỗi, mà là do chính các châm ngôn của anh ta, \& theo thiết kế chính \& đầu tiên. Bởi vì nói những điều ngớ ngẩn một cách tình cờ \& là một điều bất hạnh phổ biến; nhưng nói chúng một cách cố ý là không thể chấp nhận được; \& nói như vậy $\ldots$
		
		\fbox{63} {\it Montaigne.} -- {\sc Montaigne}'s faults are great. Lewd words; this is bad, notwithstanding {\sc Mademoiselle de Gournay}.
		
		-- {\it Montaigne.} -- Lỗi của {\sc Montaigne} rất lớn. Những lời tục tĩu; điều này thật tệ, bất chấp {\sc Mademoiselle de Gournay}.
		
		Credulous; {\it people without eyes}. Ignorant; {\it squaring the circle, a greater world}. His opinions on suicide, on death. He suggests an indifference about salvation, {\it without fear \& without repentance}. As his book was not written with a religious purpose, he was not bound to mention religion; but it is always our duty not to turn men from it.  One can excuse his rather free \& licentious opinions on some relations of life (730,231); but one cannot excuse his thoroughly pagan views on death, for a man must renounce piety altogether, if he does not at least wish to die like a Christian. Now, through the whole of his book his only conception of death is a cowardly \& effeminate one.
		
		-- Cả tin; {\it những người không có mắt}. Dốt nát; {\it làm vuông hình tròn, một thế giới vĩ đại hơn}. Quan điểm của ông về tự tử, về cái chết. Ông cho thấy sự thờ ơ về sự cứu rỗi, {\it không sợ hãi \& không ăn năn}. Vì cuốn sách của ông không được viết với mục đích tôn giáo, nên ông không bắt buộc phải đề cập đến tôn giáo; nhưng chúng ta luôn có bổn phận không được khiến con người xa lánh nó. Người ta có thể bào chữa cho những quan điểm khá tự do \& phóng túng của ông về một số mối quan hệ trong cuộc sống (730,231); nhưng người ta không thể bào chữa cho quan điểm hoàn toàn ngoại giáo của ông về cái chết, vì một người phải từ bỏ hoàn toàn lòng mộ đạo, nếu anh ta không muốn chết như một người theo đạo Thiên chúa. Bây giờ, trong toàn bộ cuốn sách của mình, quan niệm duy nhất của ông về cái chết là một quan niệm hèn nhát \& yếu đuối.
		
		\fbox{64} It is not in {\sc Montaigne}, but in myself, that I find all that I see in him.
		
		-- Không phải ở {\sc Montaigne}, mà ở chính bản thân tôi, tôi tìm thấy tất cả những gì tôi thấy ở ông.
		
		\fbox{65} What good there is in {\sc Montaigne} can only have been acquired with difficulty. The evil that is in him, I mean apart from his morality, could have been corrected in a moment, if he had been informed that he made too much of trifles \& spoke too much of himself.
		
		-- Những điều tốt đẹp ở {\sc Montaigne} chỉ có thể đạt được một cách khó khăn. Cái ác ở ông, ý tôi là ngoài đạo đức của ông, có thể được sửa chữa trong chốc lát, nếu ông được thông báo rằng ông đã làm quá nhiều chuyện vặt vãnh \& nói quá nhiều về bản thân mình.
		
		\fbox{66} One must know oneself. If this does not serve to discover truth, it at least serves as a rule of life, \& there is nothing better.
		
		-- Người ta phải biết chính mình. Nếu điều này không giúp khám phá ra chân lý, thì ít nhất nó cũng là một quy luật của cuộc sống, \& không có gì tốt hơn.
		
		\fbox{67} {\it The vanity of the sciences.} -- Physical science will not console me for the ignorance of morality in the time of affliction. But the science of ethics will always console me for the ignorance of the physical sciences.
		
		-- {\it Sự phù phiếm của khoa học.} -- Khoa học vật lý sẽ không an ủi tôi vì sự thiếu hiểu biết về đạo đức trong thời kỳ đau khổ. Nhưng khoa học đạo đức sẽ luôn an ủi tôi vì sự thiếu hiểu biết về khoa học vật lý.
		
		\fbox{68} Men are never taught to be gentlemen, \& are taught everything else; \& they never plume themselves so much on the rest of their knowledge as on knowing how to be gentlemen. They only plume themselves on knowing the 1 thing they do not know.
		
		-- Đàn ông không bao giờ được dạy để trở thành quý ông, \& được dạy mọi thứ khác; \& họ không bao giờ tự hào về phần kiến thức còn lại của mình mà là về việc biết cách trở thành quý ông. Họ chỉ tự hào về việc biết 1 điều mà họ không biết.
		
		\fbox{69} {\it The infinites, the mean.} -- When we read too fast or too slowly, we understand nothing.
		
		-- {\it Vô hạn, trung bình.} -- Khi chúng ta đọc quá nhanh hoặc quá chậm, chúng ta không hiểu gì cả.
		
		\fbox{70} {\it Nature $\ldots$} -- [Nature has set us so well in the center, that if we change 1 side of the balance, we change the other also. {\it I act.} This makes me believe that the springs in our brain are so adjusted that he who touches one touches also its contrary.]
		
		-- {\it Thiên nhiên $\ldots$} -- [Thiên nhiên đã đặt chúng ta vào vị trí trung tâm rất tốt, đến nỗi nếu chúng ta thay đổi một bên của cán cân, chúng ta cũng thay đổi bên kia. {\it Tôi hành động.} Điều này khiến tôi tin rằng các lò xo trong não chúng ta được điều chỉnh rất tốt đến mức khi ai chạm vào một bên thì cũng chạm vào bên đối diện của nó.]
		
		\fbox{71} Too much \& too little wine. Given him none, he cannot find truth; give him too much, the same.
		
		-- Quá nhiều \& quá ít rượu. Không cho anh ta, anh ta không thể tìm thấy sự thật; cho anh ta quá nhiều, cũng vậy.
		
		\fbox{72} {\it Man's disproportion.}
		
		-- {\it Sự mất cân xứng của con người.}
		
		\item {\sf Of the Necessity of the Wager.}
		\item {\sf Of the Means of Belief.}
		\item {\sf Justice \& the Reason of Effects.}
		\item {\sf The Philosophers.}
		\item {\sf Morality \& Doctrine.}
		\item {\sf The Fundamentals of the Christian Religion.}
		\item {\sf Perpetuity.}
		\item {\sf Typology.}
		\item {\sf The Prophecies.}
		\item {\sf Proofs of Jesus Christ.}
		\item {\sf The Miracles.}
		\item {\sf Appendix: Polemical Fragments.}
		\item {\sf Notes.}
	\end{itemize}	
	
	\item \cite{Pascal_pensees_other_writings}. {\sc Blaise Pascal, Honor Levi}. {\it Pens\'ees \& Other Writings}. {\sf[84 Amazon ratings][13569 Goodreads ratings]}
	
	{\sf Amazon review.} For much of his life, {\sc Pascal} (1623--62) worked on a magnum opus which was never published in the form the philosopher intended. Instead, {\sc Pascal} left a mass of fragments, some of them meant as notes for the {\it Apologie}. These became known as the {\it Pens\'ees}, \& they occupy a crucial place in Western philosophy \& religious writing.
	
	{\sc Pascal}'s general intention was to confound scepticism about metaphysical questions. Some of the {\it Pens\'ees} are fully developed literary reflections on the human condition, some contradict others, \& some remain jottings whose meaning will never be clear. The most important are among the most powerful aphorisms about human experience \& behavior ever written in any language.
	
	This translation is the only one based on the {\it Pens\'ees} as {\sc Pascal} left them. It includes the principal dossiers classified by {\sc Pascal}, as well as the essential portion of the important {\it Writings on Grace}. A detailed thematic index gives access to {\sc Pascal}'s areas of concern, while the selection of texts \& the introduction help to show why {\sc Pascal} changed the plan of his projected work before abandoning the book he might have written.
	
	{\sf About the Series.} For $>100$ years {\it Oxford World's Classics} has made available the broadest spectrum of literature from around the globe. Each affordable volume reflects Oxford's commitment to scholarship, providing the most accurate text plus a wealth of other valuable features, including expert introductions by leading authorities, voluminous notes to clarify the text, up-to-date bibliographies for further study, \& much more.
	
	A new edition of the {\it Pensées}, which occupies a crucial place in Western philosophy \& religious writing. 
	
	{\sf Oxford World's Classics.} {\sc Blaise Pascal} (1623--62), sickly as a youth \& in poor health throughout his life, was precociously gifted as a mathematician \& scientist, but is now chiefly remembered for his commitment to the religious group associated with the monastery of Port-Royal, generally if inappropriately referred to as `Jansenist'. He never wrote a book, but did compose a brilliant series of satirical \& polemical flysheets, subsequently known as the {\it Lettres provinciales}, against the relaxation of strict \& immutable moral norms, \& left a series of abandoned texts, of which the most important were post-humously published after heavy editing by his family \& admirers for purposes of religious edification. They have ever since been known after their modish 17th-century title as the {\it Pens\'ees}. Some of the texts were notes produced in the context of preparing a religious apologetic.
	
	{\sc Pascal} mixed in the upper bourgeois society of his time, was on friendly terms with important scientists, administrators, legal officials, \& some of the higher nobility. In the expression of his religious views, {\sc Pascal} brought French prose writing to new peaks of lyrical, satirical, \& polemical brilliance. Although he failed to resolve the religious problems which confronted him, he was the most brilliant religious philosopher of his century \& the author of the finest satirical work written in 17th-century France.
	
	{\sc Anthony Levi} took premature retirement as Buchanan Professor of French Language \& Literature in the University of Saint Andrews in order to undertake full-time research. His books include a 2-volume {\it Guide to French Literature}, 1992 \& 1994, \& an annotated edition of {\sc Erasmus}'s {\it Praise of Folly}.
	
	{\sc Honor Levi} read French at Warwick, discovered \& published for her Edinburgh Ph.D. the legal inventory of Richelieu's possessions at death, \& reconstituted the original layout of what is now the Palais-Royal in Paris. She taught French, published papers given in France at historical symposia, wrote on French art history \& literature, \& published a series fo encyclopedia articles on historic French towns.
	\begin{itemize}
		\item {\sf Introduction.}
		\item {\sf Note on the Text.}
		\item {\sf Select Bibliography.}
		\item {\sf A Chronology of {\sc Blaise Pascal}.}
		\item {\sf Pens\'ees.}
		\item {\sf Discussion with Monsieur de Sacy.}
		\item {\sf The Art of Persuasion.}
		\item {\sf Writings on Grace.}
		\begin{itemize}
			\item {\sf Letter on the Possibility of the Commandments.}
			\item {\sf Treatise concerning Predestination.}
		\end{itemize}
		\item {\sf Explanatory Notes.}
	\end{itemize}
\end{enumerate}

%------------------------------------------------------------------------------%

\section{Jordan Peterson. 12 Rules for Life}

\begin{quotation}
	\textit{``The most influential public intellectual in the Western world right now.''} -- New York Times
\end{quotation}

\section*{Introduction}
``\textit{12 Rules for Life: An Antidote\footnote{\textbf{antidote} [n] \textbf{1.} \textbf{antidote (to something)} a substance that controls the effects of a poison or disease; \textbf{2.} \textbf{antidote (to something)} anything that takes away the effects of something unpleasant.} to Chaos\footnote{\textbf{chaos} [n] [uncountable] a state of complete confusion \& lack of order; in physics, \textbf{chaos} is the property of a complex system whose behavior is so unpredictable that it appears random, especially because small changes in conditions can have very large effects; \textbf{chaos theory} is the branch of mathematics that deals with these complex systems.}} is a 2018 \href{https://en.wikipedia.org/wiki/Self-help_book}{self-help book} by the Canadian clinical\footnote{\textbf{clinical} [a] [only before noun] connected with the examination \& treatment of patients \& their illnesses.} psychologist\footnote{\textbf{psychologist} [n] a scientist who studies psychology.} \href{https://en.wikipedia.org/wiki/Jordan_Peterson}{Jordan Peterson}. It provides life advice through essays in abstract ethical\footnote{\textbf{ethical} [a] \textbf{1.} connected with beliefs \& principles about what is right \& wrong; \textbf{2.} morally correct or acceptable.} principles, psychology, mythology\footnote{\textbf{mythology} [n] [uncountable, countable] \textbf{1.} ancient myths in general; the ancient myths of a particular culture, society, etc.; \textbf{2.} \textbf{mythology (of something)} ideas that many people think are true but are in fact false.}, religion\footnote{\textbf{religion} [n] \textbf{1.} [uncountable] the belief in the existence of a god or gods, \& the activities that are connected with the worship of them; \textbf{2.} [countable] 1 of the systems of belief that are based on the belief in the existence of a particular god or gods.}, \& personal anecdotes\footnote{\textbf{anecdote} [n] [countable, uncountable] \textbf{1.} \textbf{anecdote (about somebody{\tt/}something)} a short, interesting or funny story about a real person or event; \textbf{2.} a personal account of an event, especially one that is considered as possibly not true or accurate.}.''[$\ldots$] ``The book is written in a more accessible style than his previous academic book, \href{https://en.wikipedia.org/wiki/Maps_of_Meaning:_The_Architecture_of_Belief}{Maps of Meaning: The Artchitecture of Belief} (1999). A sequel, \href{https://en.wikipedia.org/wiki/Beyond_Order}{Beyond Order: 12 More Rules for Life}, was published in Mar 2021.'' -- \href{https://en.wikipedia.org/wiki/12_Rules_for_Life}{Wikipedia{\tt/}12 Rules for Life}

\subsection*{Overview}

\paragraph*{Background.} ``Peterson's interest in writing the book grew out of a personal hobby of answering questions posted on \href{https://en.wikipedia.org/wiki/Quora}{Quora}; 1 such question being
\begin{question}
	\fbox{``What are the most valuable things everyone should know?'',}
\end{question}
to which his answer comprised 42 rules. The early vision \& promotion of the book aimed to include all rules, with the title ``42''. Peterson stated that it ``isn't only written for other people. It's warning to me.'''' -- \href{https://en.wikipedia.org/wiki/12_Rules_for_Life#Background}{Wikipedia{\tt/}12 Rules for Life{\tt/}overview{\tt/}background}

\paragraph*{12 Rules.} ``The book is divided into chapters with each title representing 1 of the following 12 specific rules for life as explained through an essay.
\begin{enumerate}
	\item ``Stand up straight with your shoulders back.''
	\item ``Treat yourself like you are someone you are responsible for helping.''
	\item ``Make friends with people who want the best for you.''
	\item ``Compare yourself to who you were yesterday, not to who someone else is today.''
	\item ``Do not let your children do anything that makes you dislike them.''
	\item ``Set your house in perfect order before you criticize the world.''
	\item ``Pursue what is meaningful (not what is expedient\footnote{\textbf{expedient} [n] an action that is useful or necessary for a particular purpose, but not always fair or right.}).''
	\item ``Tell the truth -- or, at least, don't lie.''
	\item ``Assume that the person you are listening to might know something you don't.''
	\item ``Be precise in your speech.''
	\item ``Do not bother children when they are skate-boarding.''
	\item ``Pet a cat when you encounter\footnote{\textbf{encounter} [v] \textbf{1.} \textbf{encounter something} to experience something, especially something unpleasant or difficult, while you are trying to do something else, \textsc{synonym}: \textbf{run into something}; \textbf{2.} \textbf{encounter something{\tt/}somebody} to discover or experience something, or meet somebody, especially something{\tt/}somebody new, unusual or unexpected, \textsc{synonym}: \textbf{come across somebody{\tt/}something}; [n] a meeting, especially one that is sudden or unexpected.} one on the street.'''' -- \href{https://en.wikipedia.org/wiki/12_Rules_for_Life#12_Rules}{Wikipedia{\tt/}12 Rules for Life{\tt/}overview{\tt/}content}
\end{enumerate} 

\paragraph*{Content.} ``The book's central idea is that ``\fbox{suffering is built into the structure of \href{https://en.wikipedia.org/wiki/Being}{being}}'' \& although it can be unbearable\footnote{\textbf{unbearable} [a] too painful, annoying or unpleasant to deal with or accept, \textsc{synonym}: \textbf{intolerable}, \textsc{opposite}: \textbf{bearable}.}, people have a choice either to withdraw\footnote{\textbf{withdraw} [v] \textbf{1.} [transitive, intransitive] (used especially about armed forces) to make people leave a place; to leave a place; \textbf{2.} [intransitive] \textbf{withdraw (to something)} to leave a room; to go away from other people; \textbf{3.} [transitive] to move something back, out or away from something; \textbf{4.} [transitive] to take money out of a bank account or financial institution; \textbf{5.} [intransitive] to stop taking part in something; \textbf{6.} [intransitive] to stop wanting to speak to, or be with, other people; \textbf{7.} [transitive] to no longer provide or offer something; to no longer make something available; \textbf{8.} [transitive] \textbf{withdraw something} to say that you no longer agree with what you said before.}, which is a ``suicidal\footnote{\textbf{suicidal} [a] (of people) very unhappy or depressed \& feeling that they want to kill themselves; (of behavior) showing this.} gesture\footnote{\textbf{gesture} [n] \textbf{1.} [countable, uncountable] \textbf{gesture (of something)} something that you do or say to show a particular feeling or intention; \textbf{2.} [countable, uncountable] a movement that you make with your hands, your head or your face to show a particular meaning.}'', or to face \& transcend\footnote{\textbf{transcend} [v] \textbf{transcend something} to be or go beyond the usual limits of something.} it. Living in a world of chaos \& order,\fbox{ everyone has ``darkness''} that can \fbox{``turn them into the monsters they're capable of being''} to satisfy their \fbox{dark impulses\footnote{\textbf{impulse} [n] \textbf{1.} [countable, usually singular, uncountable] a sudden strong wish or need to do something, without stopping to think about the results; \textbf{2.} [countable, usually singular] something that causes somebody{\tt/}something to do something or to develop \& make progress; \textbf{3.} [countable] a brief electric current, e.g. one that travels from a nerve to a muscle; \textbf{4.} [countable] (\textit{physics}) the change in momentum of an object due to a force.} in the right situations}. Scientific experiments like the \href{https://en.wikipedia.org/wiki/Inattentional_blindness#Invisible_Gorilla_Test}{Invisible Gorilla Test} show that perception\footnote{\textbf{perception} [n] \textbf{1.} [uncountable, countable] an idea, a belief or an image you have as a result of how you see or understand something; \textbf{2.} [uncountable] the way you notice things or the ability to notice things with the senses; in biology, \textbf{perception} refers to the processes in the nervous system by which a living thing becomes aware of events \& things outside itself; \textbf{3.} [uncountable] the ability to understand the true nature of something, \textsc{synonym}: \textbf{insight}.} is adjusted to aims, \& it is \fbox{better to seek \href{https://en.wikipedia.org/wiki/Meaning_(psychology)}{meaning} rather than happiness}. Peterson notes:
\begin{quotation}
	``It's all very well to think the meaning of life is happiness, but what happens when you're unhappy? Happiness is a great side effect. When it comes, accept it gratefully\footnote{\textbf{grateful} [a] \textbf{1.} feeling or showing thanks because somebody has done something kind for you or has done as you asked; \textbf{2.} used to make a request, especially in a letter or in a formal situation.}. But it's fleeting\footnote{\textbf{fleeting} [a] [usually before noun] lasting only a short time, \textsc{synonym}: \textbf{brief}.} \& unpredictable\footnote{\textbf{unpredictable} [a] that cannot be predicted because it changes a lot or depends on too many different things, \textsc{opposite}: \textbf{predictable}.}. It's not something to aim at -- because it's not an aim. \& if happiness is the purpose of life, what happens when you're unhappy? Then you're a failure.''
\end{quotation}
The book advances the idea that \fbox{people are born with an instinct\footnote{\textbf{instinct} [n] [uncountable, countable] a natural tendency for people \& animals to behave in a particular way, using the knowledge \& abilities that they were born with rather than thought or training.} for ethics \& meaning}, \& should take responsibility\footnote{\textbf{responsibility} [n] \textbf{1.} [uncountable, countable] a duty to deal with or take care of somebody{\tt/}something, so that you may be blamed if something goes wrong; \textbf{2.} [uncountable] \textbf{responsibility (for something)} blame for something bad that has happened; \textbf{3.} [countable, uncountable] a moral duty to behave well with regard to somebody{\tt/}something.} to search for meaning above their own interests (Rule 7, ``Pursue what is meaningful, not what is expedient''). Such thinking is reflected both in contemporary\footnote{\textbf{contemporary} [a] \textbf{1.} belonging to the present time, \textsc{synonym}: \textbf{modern}; \textbf{2.} (especially of people \& society) belonging to the same time as somebody{\tt/}something else; [n] a person or thing living or existing at the same time as somebody{\tt/}something else, especially somebody who is about the same age as somebody else.} stories e.g. \href{https://en.wikipedia.org/wiki/Pinocchio_(1940_film)}{Pinocchio}, \href{https://en.wikipedia.org/wiki/The_Lion_King}{The Lion King}, \& \href{https://en.wikipedia.org/wiki/Harry_Potter}{Harry Potter}, \& in ancient stories from the \href{https://en.wikipedia.org/wiki/Bible}{Bible}. To ``stand up straight with your shoulders back'' (Rule 1) is to ``accept the terrible responsibility of life'', to make self-sacrifice\footnote{\textbf{self-sacrifice} [n] [uncountable] (\textit{approving}) the act of not allowing yourself to have or do something in order to help other people.}, because the individual must rise above \href{https://en.wikipedia.org/wiki/Victimisation}{victimization}\footnote{\textbf{victimize} [v] [often passive] \textbf{victimize somebody} to make somebody suffer unfairly because you do not like them, their opinions or something that they have done.} \& ``conduct his or her life in a manner that requires the rejection\footnote{\textbf{rejection} [n] [uncountable, countable] \textbf{1.} the act of refusing to accept or consider something; \textbf{2.} the act of refusing to accept somebody for a job or position; \textbf{3.} the decision not to use, sell, publish, etc. something because its quality is not good enough; \textbf{4.} \textbf{rejection (of something)} an occasion when somebody's body does not accept a new organ after a transplant operation, by producing substances that attack the organ; \textbf{5.} the act of failing to give a person or an animal enough care or affection.} of immediate gratification\footnote{\textbf{gratification} [n] [uncountable, countable] (\textit{formal}) the state of feeling pleasure when something goes well for you or when your desires are satisfied; something that gives you pleasure, \textsc{synonym}: \textbf{satisfaction}.}, of natural \& perverse\footnote{\textbf{perverse} [a] showing a deliberate \& determined desire to behave in a way that most people think is wrong, unacceptable or unreasonable.} desires alike.'' The comparison to \href{https://en.wikipedia.org/wiki/Neurology}{neurological}\footnote{\textbf{neurological} [a] relating to nerves or to the science of neurology.} structures \& behavior of \href{https://en.wikipedia.org/wiki/Lobsters}{lobsters} is used as a natural example to the formation\footnote{\textbf{formation} [n] \textbf{1.} [uncountable] the action of forming something; the process of being formed; \textbf{2.} [countable] a thing that has been formed, especially in a particular place or in a particular way; \textbf{3.} [countable, uncountable] a particular arrangement or pattern of people or things.} of \href{https://en.wikipedia.org/wiki/Hierarchy}{social hierarchies}\footnote{\textbf{hierarchy} [n] \textbf{1.} [countable, uncountable] a system, especially in a society or an organization, in which people are organized into different levels of importance from highest to lowest; \textbf{2.} [countable] a system that ideas or beliefs can be arranged into.}.

The other parts of the work explore \& criticize the state of young men; the upbringing\footnote{\textbf{upbringing} [n] [singular, uncountable] the way in which a child is cared for \& taught how to behave while it is growing up.} that ignores \href{https://en.wikipedia.org/wiki/Sex_differences_in_humans}{sex differences} between boys \& girls (criticism of \href{https://en.wikipedia.org/wiki/Overprotective}{over-protection} \& \href{https://en.wikipedia.org/wiki/Tabula_rasa}{tabula rasa} model in \href{https://en.wikipedia.org/wiki/Social_science}{social sciences}); male-female \href{https://en.wikipedia.org/wiki/Interpersonal_relationship}{interpersonal relationships}; \href{https://en.wikipedia.org/wiki/School_shooting}{school shootings}; religion \& moral \href{https://en.wikipedia.org/wiki/Nihilism}{nihilism}\footnote{\textbf{nihilism} [n] [uncountable] (\textit{philosophy}) the belief that life has no meaning or purpose \& that religious \& moral principles have no value.}; \href{https://en.wikipedia.org/wiki/Relativism}{relativism}\footnote{\textbf{relativism} [n] [uncountable] the belief that truth is not always \& generally valid, but can be judged only in relation to other things, e.g. your personal situation.}; \& lack of respect for the values that built \href{https://en.wikipedia.org/wiki/Western_world}{Western society}.

In the last chapter, Peterson outlines the ways in which one can cope with the most tragic\footnote{\textbf{tragic} [a] \textbf{1.} making you feel very sad, usually because somebody has died or suffered a lot; \textbf{2.} [usually before noun] connected with tragedy ($=$ the style of literature).} events, events that are often \fbox{out of one's control}. In it, he describes his own personal struggle upon discovering that his daughter, Mikhaila, had a rare bone disease. The chapter is a meditation\footnote{\textbf{meditation} [n] \textbf{1.} [uncountable] the practice of thinking deeply, usually in silence, especially for religious reasons or in order to make your mind calm; \textbf{2.} [countable, usually plural] \textbf{meditation (on something)} serious thoughts on a particular subject that somebody writes down or speaks.} on how to maintain\footnote{\textbf{maintain} [v] \textbf{1.} \textbf{maintain something} to cause or enable a condition or situation to continue, \textsc{synonym}: \textbf{preserve}; \textbf{2.} \textbf{maintain something} to keep something at the same level or rate; \textbf{3.} to state strongly that something is true, even when some other people may not believe it; \textbf{4.} \textbf{maintain somebody{\tt/}something} to support somebody{\tt/}something over a long period of time by providing money, paying for food, etc.; \textbf{5.} \textbf{maintain something} to keep a building, machine, etc. in good condition by checking or repairing it regularly; \textbf{6.} \textbf{maintain a record} to write something down as a record \& keep adding the most recent information, \textsc{synonym}: \textbf{keep}.} a watchful\footnote{\textbf{watchful} [a] paying attention to what is happening in case of danger, accidents, etc.} eye on, \& cherish\footnote{\textbf{cherish} [v] (\textit{formal}) \textbf{1.} \textbf{cherish somebody{\tt/}something} to love somebody{\tt/}something very much \& want to protect them or it; \textbf{2.} \textbf{cherish something} to keep an idea, a hope or a pleasant feeling in your mind for a long time.}, life's small redeemable\footnote{\textbf{redeemable} [a] \textbf{redeemable (against something)} that can be exchanged for money or goods.} qualities (i.e., ``pet a cat when you encounter one''). It also outlines a practical way to deal with hardship\footnote{\textbf{hardship} [n] [uncountable, countable] a situation that is difficult \& unpleasant because you do not have enough money, food, clothes, etc.}: to shorten one's temporal\footnote{\textbf{temporal} [a] \textbf{1.} connected with or limited by time; \textbf{2.} connected with the real physical world, not spiritual matters; \textbf{3.} (\textit{anatomy}) near the temples at the side of the head.} scope of responsibility (e.g., focusing on the next minute rather than the next 3 months).

Canadian psychiatrist \& psychoanalyst \href{https://en.wikipedia.org/wiki/Norman_Doidge}{Norman Doidge} wrote \cite{Peterson2018}'s foreword.'' -- \href{https://en.wikipedia.org/wiki/12_Rules_for_Life#Content}{Wikipedia{\tt/}12 Rules for Life{\tt/}overview{\tt/}content}

\begin{quotation}
	``The most influential public intellectual\footnote{\textbf{intellectual} [a] [usually before noun] connected with or using a person's ability to think in a logical way \& understand things, \textsc{synonym}: \textbf{mental}; [n] a person who is well educated \& enjoys activities in which they have to think seriously about things.} in the Western world right now.'' -- New York Times
\end{quotation}

%------------------------------------------------------------------------------%

\subsection{Foreword by Noman Doidge}
``Rules? More rules? Really? Isn't life complicated enough, restricting enough, without abstract rules that don't take our unique, individual situations into account? \& given that our brains are plastic, \& all develop differently based on our life experiences, why even expect that a few rules might be helpful to us all?

People don't clamor for rules, even in the Bible $\ldots$ as when Moses comes down the mountain, after a long absence, bearing the tablets inscribed with 10 commandments, \& finds the Children of Israel in revelry. They'd been Pharaoh's slaves \& subject to his tyrannical regulations for 400 years, \& after that Moses subjected them to the harsh desert wilderness for another 40 years, to purify them of their slavishness. Now, free at last, they are unbridled, \& have lost all control as they dance wildly around an idol, a golden calf, displaying all manner of corporeal corruption.

``I've got some good news $\ldots$ \& I've got some bad news,'' the lawgiver yells to them. ``Which do you want 1st?''

``The good news!'' the hedonists reply.

``I got Him from 15 commandments down to 10!''

``Hallelujah!'' cries the unruly crowd. ``\& the bad?''

``Adultery is still in.''

So rules there will be -- but, please, not too many. We are ambivalent about rules, even when we know they are good for us. If we are spirited souls, if we have character, rules seem restrictive, an affront to our sense of agency \& our pride in working out our own lives. Why should we be judged according to another's rule?

\& judged we are. After all, God didn't give Moses ``The 10 Suggestions,'' he gave Commandments; \& if I'm a free agent, my 1st reaction to a command might just be that nobody, not even God, tells me what to do, even if it's good for me. But the story of the golden calf also reminds us that without rules we quickly become slaves to our passions -- \& there's nothing freeing about that.

\& the story suggests something more: unchaperoned, \& left to our own untutored judgment, we are quick to aim low \& worship qualities that are beneath us -- in this case, an artificial animal that brings out our own animal instincts in a completely unregulated way. The old Hebrew story makes it clear how the ancients felt about our prospects for civilized behavior in the absence of rules that seek to elevate our gaze \& raise our standards.

1 neat thing about the Bible story is that it doesn't simply list its rules, as lawyers or legislators or administrators might; it embeds them in a dramatic tale that illustrates why we need them, thereby making them easier to understand. Similarly, in this book Prof. Peterson doesn't just propose his 12 rules, he tells stories, too, bringing to bear his knowledge of many fields as he illustrates \& explains why the best rules do not ultimately restrict us but instead facilitate our goals \& make for fuller, freer lives. p. 6

'' -- \cite[pp. 5--]{Peterson2018}



%------------------------------------------------------------------------------%

\subsection{Overture}

%------------------------------------------------------------------------------%

\subsection{Rule 1: Stand Up Straight with Your Shoulders Back}

%------------------------------------------------------------------------------%

\subsection{Rule 2: Treat Yourself Like Someone You Are Responsible for Helping}

%------------------------------------------------------------------------------%

\subsection{Rule 3: Make Friends with People Who Want The Best for You}

%------------------------------------------------------------------------------%

\subsection{Rule 4: Compare Yourself to Who You Were Yesterday, Not to Who Someone Else Is Today}

%------------------------------------------------------------------------------%

\subsection{Rule 5: Do Not Let Your Children Do Anything That Makes You Dislike Them}

%------------------------------------------------------------------------------%

\subsection{Rule 6: Set Your House In Perfect Order Before You Criticize The World}

%------------------------------------------------------------------------------%

\subsection{Rule 7: Pursue What Is Meaningful (Not What Is Expedient)}

\subsubsection{Get While The Getting's Good}
``Life is suffering. That's clear. There is no more basic, irrefutable truth. It's basically what God tells Adam \& Eve, immediately before he kicks them out of Paradise.
\begin{quotation}
	Unto the woman he said, I will greatly multiply thy sorrow \& thy conception; in sorrow thou shalt bring 4th children; \& thy desire shall be to thy husband, \& he shall rule over thee.
	
	\& unto Adam he said, Because thou hast hearkened unto the voice of thy wife, \& hast eaten of the tree, of which I commanded thee, saying, Thou shalt not eat of it: cursed is the ground for thy sake; in sorrow shalt thou eat of it all the days of thy life;
	
	Thorns also \& thistles shall it bring 4th to thee; \& thou shalt eat the herb of the field;
	
	By the sweat of your brow you will eat your food until you return to the ground, since from it you were taken; for dust you are \& to dust you will return.'' (Genesis 3:16-19. KJV)
\end{quotation}
What in the world should be done about that?

The simplest, most obvious, \& most direct answer? Pursue pleasure. Follow your impulses. Live for the moment. Do what's expedient. Lie, cheat, steal, deceive, manipulate -- but don't get caught. In an ultimately meaningless universe, what possible difference could it make? \& this is by no means a new idea. The fact of life's tragedy \& the suffering that is part of it has been used to justify the pursuit of immediate selfish gratification for a very long time.
\begin{quotation}
	Short \& sorrowful is our life, \& there is no remedy when a man comes to his end, \& no one has been known to return from Hades.
	
	Because we were born by mere chance, \& hereafter we shall be as though we had never been; because the breath in our nostrils is smoke, \& reason is a spark kindled by the beating of our hearts.
	
	When it is extinguished, the body will turn to ashes, \& the spirit will dissolve like empty air. Our name will be forgotten in time \& no one will remember our works; our life will pass away like the traces of a cloud, \& be scattered like mist that is chased by the rays of the sun \& overcome by its heat.
	
	For our allotted time is the passing of a shadow, \& there is no return from our death, because it is sealed up \& no one turns back.
	
	Come, therefore, let us enjoy the good things that exist, \& make use of the creation to the full as in youth.
	
	Let us take our fill of costly wine \& perfumes, \& let no flower of spring pass by us.
	
	Let us crown ourselves with rosebuds before they wither.
	
	Let none of us fail to share in our revelry, everywhere let us leave signs of enjoyment, because this is our portion, \& this our lot.
	
	Let us oppress the righteous poor man; let us not spare the widow nor regard the gray hairs of the aged.
	
	But let our might be our law of right, for what is weak proves itself to be useless.
	
	(Wisdom 2:1-11, RSV).
\end{quotation}
The pleasure of expediency may be fleeing, but it's pleasure, nonetheless, \& that's something to stack up against the terror \& pain of existence. Every man for himself, \& the devil take the hindmost, as the old proverb has it. Why not simply take everything you can get, whenever the opportunity arises? Why not determine to live in that manner?

Or is there an alternative, more powerful \& more compelling?

Our ancestors worked out very sophisticated answers to such questions, but we still don't understand them very well. This is because they are in large part still implicit -- manifest primarily in ritual \& myth \&, as of yet, incompletely articulated. We act them out \& represent them in stories, but we're not yet wise enough to formulate them explicitly. We're still chimps in a troupe, or wolves in a pack. we know how to behave. We know who's who, \& why. We've learned that through experience. Our knowledge has been shaped by our interaction with others. We've established predictable routines \& patterns of behavior -- but we don't really understand them, or know where they originated. They've evolved over great expanses of time. No one was formulating them explicitly (at least not in the dimmest reaches of the past), even though we've been telling each other how to act forever. 1 day, however, not so long ago, we woke up. We were already doing, but we started \textit{noticing} what we were doing. We started using our bodies as devices to represent their own actions. We started imitating \& dramatizing. We invented ritual. We started acting out our own experiences. Then we started to tell stories. We coded our observations of our own drama in these stories. In this manner, the information that was 1st only embedded in our behavior became represented in our stories. But we didn't \& still don't understand what it all means.

The Biblical narrative of Paradise \& the Fall is 1 such story, fabricated by our collective imagination, working over the centuries. It provides a profound account of the nature of Being, \& points the way to a mode of conceptualization \& action well-matched to that nature. In the Garden of Eden, prior to the dawn of self-consciousness -- so goes the story -- human beings were sinless. Our primordial parents, Adam \& Eve, walked with God. Then, tempted by the snake, the 1st couple ate from the tree of the knowledge of good \& evil, discovered Death \& vulnerability, \& turned away from God. Mankind was exiled from Paradise, \& began its effortful mortal existence. The idea of sacrifice enters soon afterward, beginning with the account of Cain \& Abel, \& developing through the Abrahamic adventures \& the Exodus: After much contemplation, struggling humanity learns that God's favor could be gained, \& his wrath averted, through proper sacrifice -- \&, also, that bloody murder might be motivated among those unwilling or unable to succeed in this manner.'' -- \cite[pp. 183--185]{Peterson2018}

\subsubsection{The Delay of Gratification}

%------------------------------------------------------------------------------%

\subsection{Rule 8: Tell The Truth -- Or, At Least, Don't Lie}

%------------------------------------------------------------------------------%

\subsection{Rule 9: Assume That The Person You Are Listening to Might Know Something You Don't}

%------------------------------------------------------------------------------%

\subsection{Rule 10: Be Precise In Your Speech}

%------------------------------------------------------------------------------%

\subsection{Rule 11: Do Not Bother Children When They Are Skateboarding}

%------------------------------------------------------------------------------%

\subsection{Rule 12: Pet A Cat When You Encounter One On The Street}

%------------------------------------------------------------------------------%

\subsection{Coda}

%------------------------------------------------------------------------------%

\subsection{Endnotes}

%------------------------------------------------------------------------------%

\section{Jordan Peterson. Beyond Order}

%------------------------------------------------------------------------------%

\section{Meaning of Life -- Giá Trị Của Sự Sống}

%------------------------------------------------------------------------------%

\section{Meaning of Death -- Giá Trị Của Cái Chết}

%------------------------------------------------------------------------------%

\section{Miscellaneous}

%------------------------------------------------------------------------------%

\printbibliography[heading=bibintoc]
	
\end{document}